---
title: "Non-negative, Approximately Orthogonal Bases: Methodology and Experiments"
subtitle: "NSA-Flow v2 --- one functional, one parameter, one solver"
author: "Research Draft"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      cache = FALSE, fig.width = 9, fig.height = 4.5)
library(ggplot2); library(dplyr); library(tidyr); library(knitr); library(reticulate)
Sys.setenv(PYTHONPATH = normalizePath(file.path("..")))
sup  <- import("experiments.rmd_support")
nsa  <- import("nsa_flow")
np   <- import("numpy")
skp  <- import("sklearn.preprocessing")
dat  <- import("experiments.data")
theme_set(theme_minimal(base_size = 11) +
          theme(panel.grid.minor = element_blank(), legend.position = "bottom"))
PAL <- c("#1f78b4", "#d9860a", "#33a02c", "#c0392b", "#7b3294", "#888888")
# reticulate may or may not have already converted the pandas frame
py_df <- function(x) {
  if (inherits(x, "data.frame")) return(as.data.frame(x))
  as.data.frame(py_to_r(x))
}
```

# Abstract

We study the problem of approximating a data matrix by a basis that is
non-negative and *approximately* orthogonal, with the degree of approximation
controlled by a single interpretable parameter. The contribution is a
consolidation rather than an addition: one energy in place of a family of
retraction schemes, one convex weight in place of several calibration constants,
and one solver with a computable stationarity certificate. We give the
functional that makes this well posed, show that its zero set under
non-negativity is exactly a hard clustering of features, and report experiments
that are deliberately unflattering --- including two places where the method
does not beat the obvious baseline, and one widely-used preprocessing step that
we show destroys three quarters of the achievable reconstruction.

# Introduction

A loading matrix that is non-negative and orthogonal is interpretable in a
specific and useful sense: non-negativity means the entries are magnitudes, and
orthogonality under non-negativity forces the columns onto *disjoint feature
supports*, so each component names a set of features and no feature is shared.
The difficulty is that both conditions together are very rigid. Exactly disjoint
supports are a combinatorial object; insisting on them turns a smooth problem
into a clustering problem and discards the data's ability to say that two
features genuinely belong together.

The method studied here relaxes one of the two. Non-negativity is kept as a hard
constraint; orthogonality becomes an objective with weight $w \in [0,1]$, so that
$w \to 0$ returns the unconstrained fit and $w \to 1$ returns the clustering.
The interesting regime is the interior, and the practical claim is that the
interior is reachable, cheap, and controlled by one number.

Three things distinguish this version from the earlier retraction-based
formulation. The orthogonality functional is *complete*, in a sense made precise
in Section \@ref(functional). There is no factorisation in the solver loop. And
the parameter needs no calibration, because both terms of the energy are
dimensionless and share their endpoints.

# Methods

## The functional {#functional}

For $Y \in \mathbb{R}^{p\times k}$ write $G = Y^{\mathsf T}Y / \operatorname{tr}(Y^{\mathsf T}Y)$
for the trace-normalised Gram matrix. Two candidate measures of "how far from
orthogonal" are

$$D(Y) = \Big\lVert G - \tfrac{1}{k}I\Big\rVert_F^2, \qquad
  C(Y) = \frac{1}{k(k-1)}\sum_{i \ne j}\cos^2(y_i, y_j),$$

with $\cos(y_i,y_j) = \langle y_i,y_j\rangle/(\lVert y_i\rVert\lVert y_j\rVert)$.
These differ in a way that matters. $D$ decomposes as

$$D = \underbrace{\sum_{i\ne j} G_{ij}^2}_{\text{decorrelation}} \;+\;
      \underbrace{\sum_i \big(G_{ii} - \tfrac1k\big)^2}_{\text{norm balance}},$$

so $D = 0$ requires both mutual orthogonality **and** equal column norms --- that
is orthoNORMality. $C = 0$ requires only orthogonality, at any norms. The
earlier formulation used the decorrelation term alone, which is weaker than
either: a zero column is orthogonal to everything, so that functional is
minimised by rank collapse.

Which of the three is right depends on what is being asked for, and the
experiment below is the cleanest way to see it.

```{r properties}
props <- py_df(sup$properties_table())
kable(props, digits = 5, col.names = c("case", "D (orthonormality)",
      "d (v1, off-diagonal)", "C (orthogonality)"),
      caption = "Each functional evaluated on the structures that separate them. Zero means 'this is what I am looking for'.")
```

Read the second and third rows. A basis with orthogonal columns of unequal norms,
and a perfectly disjoint non-negative basis whose components differ in magnitude,
are both exactly what this method exists to produce --- and $D$ charges them
$0.69$ and $0.14$ respectively, because their column norms are not equal. Real
components do differ in magnitude, so $D$ penalises the target. Read the fourth
row for the opposite failure: the v1 functional scores a rank-collapsed matrix at
exactly $0$. $C$ is zero on the first three and positive on the last two, which
is the behaviour wanted.

We therefore use $C$ when orthogonality is the goal, and note that the two
coincide in the data-anchored setting of Section \@ref(anchors) because the
reconstruction term already forces the norms to match.

## What $w \to 1$ means

The relaxation is not a convenience. For $Y \ge 0$, $Y^{\mathsf T}Y$ is diagonal
if and only if the columns have pairwise disjoint supports, so the feasible set
at $w = 1$ is a set of hard assignments of features to components --- equivalently
orthogonal non-negative matrix factorisation, which is $k$-means on the features.
That set is a union of lower-dimensional strata, connected only through
degenerate configurations. Optimising over it directly is a combinatorial
problem; the relaxation is what makes gradient information usable, and $w$ is how
far toward that limit one chooses to go.

## Two anchors {#anchors}

The same orthogonality term supports two different problems, and the distinction
turned out to matter more than anything else in this study.

**Refinement.** Given a loading matrix $X_0$ that is already trusted, find a
nearby non-negative, near-orthogonal one:

$$\min_{Y \ge 0}\; (1-w)\,\frac{\lVert Y - X_0\rVert_F^2}{\lVert X_0\rVert_F^2}
  \;+\; w\,\tilde D(Y).$$

**Estimation.** Given the data $X \in \mathbb{R}^{n \times p}$, find the basis
directly:

$$\min_{V \ge 0}\; (1-w)\,\frac{\lVert X - XVV^{\mathsf T}\rVert_F^2}{\lVert X\rVert_F^2}
  \;+\; w\,C(V).$$

Scores are *determined* by projection, $U = XV$, rather than being a second free
block, so $V$ remains the only variable: one smooth objective on a convex set,
and an out-of-sample extension $X_{\text{new}}V$.

Both terms are dimensionless and share their endpoints --- each is $0$ at its own
ideal and exactly $1$ at a natural reference ($V = 0$ for the fidelity, rank one
for $\tilde D$) --- which is why $w$ requires no calibration constants.

A further convenience: $XVV^{\mathsf T}$ is an orthogonal projection exactly when
$V^{\mathsf T}V = I$, so the reconstruction term already prefers orthonormal
columns. Fidelity and orthogonality cooperate here rather than trading off.

## Relaxing into non-negativity, rather than projecting into it

The refinement form needs a non-negative starting point, and the obvious
construction is wrong. Taking $\lvert L \rvert$ of a signed basis is not the
projection onto $\{V \ge 0\}$ --- for an entry $-3$ it moves distance $6$ where
the projection moves $3$ --- and it has no variational characterisation at all.
More fundamentally, *no* one-shot map can work: a signed component encodes a
contrast, and a single non-negative column cannot represent a contrast.

We therefore follow a path. With $\mu$ increasing from $0$,

$$\min_V\; E_w(V) \;+\; \mu\,\lVert \min(0,V)\rVert_F^2,$$

each subproblem smooth, each warm-started from the last. At $\mu = 0$ the
minimiser is the signed basis --- the global optimum of an easy problem --- and
increasing $\mu$ deforms it continuously into the feasible set.

## Solver

Spectral projected gradient: Barzilai--Borwein step lengths with Armijo
backtracking on the projected step. The gradient-mapping norm
$\lVert Y^+ - Y\rVert/t$ is a computable stationarity certificate, and the solver
reports which stopping rule fired rather than asserting convergence.

No SVD, eigendecomposition or QR appears in the loop. For $p > n$ the Gram matrix
is never formed either: every quantity follows from $XV$ and $X^{\mathsf T}(XV)$,
giving $O(npk)$ per iteration.

# Experiments

All experiments run from the installed package. Real-data fits are
cross-validated with every step --- scaling, covariate adjustment, basis,
classifier --- estimated inside the training fold.

```{r load-adni}
adni <- dat$load_adni("right")
X_raw <- py_to_r(adni[[1]]); meta <- py_to_r(adni[[2]])
cov_mat <- cbind(1, meta$AGE, as.numeric(as.character(meta$SEX) == "M"))
beta <- qr.solve(cov_mat, X_raw)
Xa <- scale(X_raw - cov_mat %*% beta)
cat(sprintf("ADNI right hemisphere: n = %d subjects, p = %d regions\n",
            nrow(Xa), ncol(Xa)))
```

## The cost of a non-negative basis is not what it appears

```{r abs-trap}
at <- py_df(sup$abs_trap_table(np$array(Xa)))
kable(at, digits = 4, caption = "Relative reconstruction error of a rank-5 basis, four ways.")
```

This is the single most consequential measurement in the study. Taking
`abs()` of the PCA loadings --- a common way to obtain a non-negative basis, and
the one the earlier version of this work used --- costs a factor of
`r round(at$cost_vs_signed[2], 2)` in reconstruction error. The correct
projection is better but still costs `r round(at$cost_vs_signed[3], 2)`.
Fitting a non-negative basis *to the data* costs
`r round(at$cost_vs_signed[4], 3)`.

Non-negativity is therefore nearly free; what is expensive is obtaining it by
mapping a signed basis instead of by fitting. Any accuracy penalty previously
attributed to the non-negativity constraint on this data was mostly attributable
to the `abs()`.

## One parameter, monotone in what it controls

```{r dial, fig.height=3.6}
dl <- py_df(sup$w_dial_table(np$array(Xa)))
kable(dl, digits = 5, caption = "Sweeping w in the data-anchored form.")
dl %>%
  select(w, `reconstruction error` = recon_err, `support overlap` = overlap,
         `orthogonality defect C` = C) %>%
  pivot_longer(-w) %>%
  ggplot(aes(w, value)) + geom_line(colour = PAL[1], linewidth = 0.9) +
  geom_point(colour = PAL[1], size = 1.6) +
  facet_wrap(~name, scales = "free_y") +
  labs(x = "w", y = NULL, title = "w trades reconstruction for disjointness, monotonically")
```

Over the whole range the orthogonality defect falls by roughly three orders of
magnitude and support overlap by `r round(dl$overlap[1] / dl$overlap[nrow(dl)], 1)`
times, at a reconstruction cost of
`r sprintf("%.1f%%", 100 * (dl$recon_err[nrow(dl)] / dl$recon_err[1] - 1))`.
Every setting converged on its own certificate.

## The relaxation path, not the iteration budget, decides the answer

```{r homotopy}
hm <- py_df(sup$homotopy_table(np$array(Xa)))
kable(hm, digits = 5, caption = "Final objective under coarse and fine mu schedules, at two iteration budgets.")
```

Refining the path from one stage to nine improves the attained objective by
roughly a factor of two. Increasing the per-stage iteration budget twenty-fold
changes it in the fourth decimal. This is worth stating plainly because the fine
schedule costs about nine times as much work and is therefore tempting to trim:
trimming it does not cost time, it costs the answer.

## Reproducibility of the basis

```{r spectrum}
sp <- py_df(sup$adni_spectrum(6L))
kable(sp, digits = 3, caption = "Leading eigenvalues of the analysis matrix and consecutive gap ratios.")
```

Component 1 is well separated; components 2--5 have gap ratios between
`r sprintf("%.2f", min(sp$gap_ratio[2:5]))` and
`r sprintf("%.2f", max(sp$gap_ratio[2:5]))`. By the Davis--Kahan theorem the error
of an individual eigenvector scales with the inverse eigengap, so PCA's
components 2--5 have poorly determined orientation on this data and should move
between resamples.

```{r repro}
rp <- py_df(sup$reproducibility_table())
kable(rp, digits = 5, caption = "Matched |cos| between bases fitted on independent folds; 10 fold pairs.")
```

Both constrained methods are more reproducible than PCA, and sparse PCA is the
most reproducible of the three. We had earlier reported this as an advantage
specific to the method; on the wider evidence that claim does not hold, and the
honest statement is that a sparsity or disjointness constraint selects a stable
representative from a degenerate subspace, whichever constraint is used.

## Accuracy against interpretability on real data

```{r frontier, fig.height=4.2}
fr <- py_df(sup$adni_frontier(n_repeats = 2L))
kable(fr, digits = 4, caption = "Cross-validated AUC and support overlap, ADNI, k = 5.")
ggplot(fr, aes(overlap, auc, colour = family, shape = family)) +
  geom_point(size = 2.6) + facet_wrap(~task, scales = "free_y") +
  scale_colour_manual(values = PAL) +
  labs(x = "support overlap (extra components per active feature; lower is more interpretable)",
       y = "cross-validated AUC",
       title = "The trade-off, read left-to-right: how much accuracy does disjointness cost?")
```

The honest reading: at matched overlap the method is level with sparse PCA and
slightly below PCA, and it does not overtake either. What it does deliver is
overlap well below what sparse PCA reaches at comparable accuracy, obtained from
one parameter rather than a penalty that must be tuned per dataset.

## Cost

```{r scaling}
sc <- py_df(sup$scaling_table())
kable(sc, digits = 4, caption = "Seconds for a fixed iteration budget, forming the Gram matrix against matrix-free.")
```

The Gram matrix is $p \times p$, so it is the binding constraint long before the
arithmetic is: at $p = 8000$ it alone is
`r sprintf("%.1f GB", sc$gram_GB[nrow(sc)])`. Working from $X$ instead makes each
iteration $O(npk)$ and removes that allocation entirely.

# Discussion

The consolidation is the contribution. One energy replaces a family of retraction
schemes; one convex weight replaces several calibration constants; one solver
carries a certificate and no factorisation. The functional is now the completed
object rather than one half of it, and the choice between $D$ and $C$ is settled
by what is being asked for rather than by convention.

Two results deserve emphasis because they are corrections rather than additions.
The `abs()` measurement shows that a widely-used preprocessing step, not the
non-negativity constraint, accounts for most of the reconstruction penalty
usually attributed to non-negative methods on this kind of data. And the homotopy
result shows that for this objective the relaxation schedule is part of the
method, not a tuning detail: a coarse schedule converges just as tightly to a
materially worse answer.

# Limitations

We state these as measured, not as caveats.

**It does not improve prediction.** Across ADNI, a Parkinson's cohort with a
biomarker-defined label, a high-dimensional expression dataset and several public
benchmarks, the method is level with sparse PCA and slightly below PCA at matched
interpretability. It is not a better predictor and should not be presented as one.

**The reproducibility advantage is not structural.** It holds against PCA on this
data and on some others, but sparse PCA is frequently better, and on two of six
datasets tested the method was clearly worse.

**Small $p$ relative to $k$ is unfavourable.** Near-disjoint supports need room:
with $k$ components in $p$ features each gets about $p/k$, and where that is small
the constraint is costly. The largest losses we measured were at $p = 13$ and
$p = 66$.

**The objective is non-convex.** The reconstruction term is quartic in $V$, so the
solver returns a stationary point. The homotopy is what makes the outcome
reproducible in practice, and its schedule must not be trimmed.

**A signed lifting does not resolve the trade-off.** Writing $V = V^+ - V^-$
restores a signed basis's reconstruction exactly, confirming the non-negative
ceiling is representational. But it reproduces PCA's density along with PCA's
accuracy, and once forced to be sparse it does worse than the plain form,
because $2k$ disjoint parts need twice the room.

# Appendix: environment

```{r env}
cat(sprintf("nsa_flow %s\n", nsa$`__version__`))
cat(sprintf("torch %s | numpy %s\n", import("torch")$`__version__`, np$`__version__`))
cat(R.version.string, "\n")
```
