---
title: "Non-Negative Stiefel Approximating Flow: Orthogonalish Matrix Optimization for Interpretable Embeddings"
author: |
  Brian B. Avants, Nicholas J. Tustison, and James R. Stone  
  Department of Radiology and Medical Imaging  
  University of Virginia, Charlottesville, VA 22903
date: "October 11, 2025"
output:
  bookdown::pdf_document2:
    keep_tex: true
    latex_engine: xelatex
    number_sections: true
    toc: false
header-includes:
  - \newcommand{\diag}{\operatorname{diag}}
  - \newcommand{\fid}{\operatorname{fid}}
  - \newcommand{\orth}{\operatorname{orth}}
  - \newcommand{\sym}{\operatorname{sym}}
  - \newcommand{\sign}{\operatorname{sign}}
  - \newcommand{\tol}{\operatorname{tol}}
  - \newcommand{\mean}{\operatorname{mean}}
  - \newcommand{\cand}{\operatorname{cand}}
link-citations: yes
linkcolor: blue
urlcolor: blue
bibliography: orth.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE,  warning = FALSE, fig.align = 'center', fig.width = 6, fig.height = 4)
set.seed(42)
library(ggplot2); library(gridExtra); library(reshape2); library(pheatmap); library(RColorBrewer)
library(dplyr); library(tidyr); library(scales); library(knitr); library(DiagrammeR)
options(width = 120)
library(ANTsR)          # retained for nsa_flow_flowchart and imaging helpers only

## The NSA-Flow solver now comes from the Python package rather than from
## ANTsR, whose implementation is the superseded v1.  nsa_shim.R keeps this
## document's original call signatures and routes them to the current code;
## see the shim header for what each v1 argument now maps to, and
## nsa_shim_ignored() for the v1-only arguments that no longer have a referent.
##
## LOAD ORDER MATTERS.  reticulate initialises PyTorch, which brings its own
## OpenMP/BLAS runtime; loading caret or randomForest BEFORE it segfaults this
## document at the first solver call.  Initialise Python first, then the R
## modelling packages.
source("nsa_shim.R")
NSA_VERSION <- nsa_shim_init(normalizePath(file.path("..")))
library(reticulate); library(tidyr); library(kableExtra)
sup <- import("experiments.rmd_support")

## Previously loaded from inside the complexity-table chunk, which is a poor
## place for them: that chunk looks like it only draws a table, and removing it
## broke the Golub analysis forty chunks later.
library(caret)          # trainControl / train, used by the Golub analysis
library(Rtsne)          # t-SNE embedding for the subtype figure
library(randomForest)   # the classifier for the diagnosis analyses
library(gt)             # the diagnosis stability table (tab_style, fmt_number)

## One table style for the whole document.  booktabs rules rather than
## full grids, no extra vertical space every five rows, numeric columns right
## aligned and text left, and the float held where it is written.
ktab <- function(x, caption, digits = 3, align = NULL, ...) {
  if (is.null(align)) {
    align <- vapply(x, function(col)
      if (is.numeric(col)) "r" else "l", character(1))
  }
  knitr::kable(x, caption = caption, digits = digits, booktabs = TRUE,
               linesep = "", align = align, ...) |>
    kableExtra::kable_styling(latex_options = c("hold_position"),
                              font_size = 9, position = "center")
}
py_to_r_df <- function(x) if (inherits(x, "data.frame")) as.data.frame(x) else as.data.frame(reticulate::py_to_r(x))
```


# Abstract {-}

NSA-Flow estimates a matrix that is non-negative and approximately orthogonal, with the degree of approximation set by one parameter. For a non-negative matrix, orthogonal columns have disjoint supports, so each component names a set of features and no feature is shared. Exact disjointness is a clustering of the features and discards the data's ability to say that two features belong together; the relaxation is what makes the interior of that constraint reachable. We keep non-negativity as a hard constraint and treat orthogonality as an objective weighted by \( w \in [0,1] \), so \( w = 0 \) returns the unconstrained fit and \( w = 1 \) returns the clustering. Both terms of the energy are dimensionless, so \( w \) needs no calibration. The solver is a spectral projected gradient with a computable stationarity certificate and no factorisation in the loop, at \( O(pk^2) \) per iteration, or \( O(npk) \) when the basis is fitted to data with \( p > n \). NSA-Flow can refine a supplied basis, fit one to data directly, or act as the proximal step inside another algorithm. On the Golub leukemia data it raises cross-validated accuracy from \( 0.828 \) to \( 0.875 \) as the sparsification step in sparse PCA. On ADNI cortical thickness it raises out-of-sample \( \Delta R^2 \) for nine cognitive outcomes by \( 0.043 \) over PCA and trinary diagnosis AUC from \( 0.719 \) to \( 0.765 \). We also show that comparisons of this kind must use a model that is sensitive to the choice of axes: a linear model on projected scores is exactly invariant to reparametrisation of the basis and cannot distinguish two bases spanning the same subspace.

# Introduction

Data matrices in neuroscience, genomics and text encode overlapping sources of variation: voxels carry distributed activity, genes are co-expressed across pathways, words co-occur across topics. A basis fitted to such data spreads each component across most features, which makes the components hard to name. Two constraints make them nameable. Non-negativity turns loadings into magnitudes. Orthogonality, imposed on a non-negative matrix, forces the columns onto disjoint feature supports. Taken exactly, the pair is very rigid: disjoint supports are a hard assignment of features to components, so the constrained problem is a clustering and the data can no longer say that two features belong together. This paper is about the interior of that constraint, and about reaching it with one parameter.

Classical dimensionality reduction techniques like principal component analysis (PCA) and its sparse variants seek low-rank approximations with interpretable bases [@zou2006sparse], while sparse canonical correlation analysis (CCA) extends this to multi-view correlations in biological data [@witten2011]. However, enforcing sparsity and decorrelation remains challenging: traditional methods may over-regularize or lack intuitive controls for partial constraints. Non-negative matrix factorization (NMF) [@lee2001nmf] offers parts-based, additive representations aligned with domain constraints, but suffers from rotational ambiguity, yielding entangled factors [@ding2006orthogonal]. Orthogonal variants improve sparsity and identifiability by aligning factors with disjoint structures, with applications in sparse PCA [@zou2006sparse], sparse CCA [@witten2011], and interpretable neural networks [@henaff2011deep]. Yet, strict orthogonality often sacrifices fidelity, especially in noisy or heterogeneous data and motivates the need for partial decorrelation models.

Soft orthogonalization methods address this by penalizing deviations from orthonormality. Examples include Disentangled Orthogonality Regularization (DOR), which separates Gram matrix components for convolutional kernels [@wu2023towards]; Group Orthogonalization Regularization (GOR), which applies intra-group penalties for vision tasks [@kurtz2023group]; $\lambda$-Orthogonality Regularization, which introduces thresholded penalties for representation learning [@ricci2025orthogonality]; and simpler approaches like Spectral Restricted Isometry Property (SRIP) [@goessmann2020restricted] and Frobenius norm penalties for neural stability [@guo2019frobenius]. However, these methods are typically embedded in neural training pipelines and do not enforce non-negativity, limiting their applicability to domains where measurement units should be preserved. Advanced ONMF variants, such as variational Bayesian approaches [@rahiche2022variational], unilateral factorization [@li2023unilateral], and deep autoencoder frameworks [@yang2021orthogonal], improve robustness but enforce strict orthogonality or require full decomposition, reducing flexibility for one-sided refinement.

In contrast to these soft regularization techniques, Riemannian optimization approaches have been explored primarily for enforcing strict orthogonality constraints on manifolds such as the Stiefel manifold, where the feasible set is equipped with a differential structure for gradient-based updates. For instance, Nonlinear Riemannian Conjugate Gradient (NRCG) methods optimize orthogonal NMF by projecting gradients onto the tangent space and using retractions like QR decomposition to maintain exact orthonormality, while handling non-negativity through coordinate descent on the complementary factor [@Zhang2016EfficientON]. This ensures convergence to critical points with near-perfect orthogonality but incurs higher computational costs compared to soft penalties, and it typically requires full enforcement rather than flexible deviations. Hybrid methods like Feedback Gradient Descent (FGD) attempt to bridge this gap by approximating manifold dynamics in Euclidean space with feedback terms to achieve stable near-orthogonality efficiently, outperforming traditional Riemannian methods in DNN training speed while rivaling soft constraints in overhead [@Bu2022FeedbackGD]. However, adapting such Riemannian-inspired techniques to incorporate soft orthogonality penalties alongside non-negativity remains a less explored avenue.

To address these gaps we propose **Non-negative Stiefel Approximating Flow (NSA-Flow)**, which approximates a target \( X_0 \) with a non-negative matrix \( Y \in \mathbb{R}^{p \times k}_{\geq 0} \) whose columns are approximately orthogonal. One weight \( w \in [0,1] \) sets the balance between fidelity and orthogonality, and because both energy terms are dimensionless it needs no calibration. Non-negativity is a hard constraint applied by exact projection; orthogonality is an objective, so nothing is retracted and no factorisation appears in the iteration. For a non-negative matrix, orthogonal columns have disjoint supports, so \( w \) moves the solution between the unconstrained fit and a hard assignment of features to components. NSA-Flow can refine a basis that is already trusted, fit one to data directly, or serve as the proximal step inside another algorithm; the Golub result below is of the third kind.

Our contributions are:

1. A general framework for constrained matrix approximation, parameterized by ( $w$ ) to intuitively control sparsity and orthogonality for interpretable ML.

2. Rigorous empirical validation that demonstrates good convergence properties and reliable benchmark performance compared to baselines.

3. Broad applications, including enhanced disease classification on the Golub leukemia dataset, non-negative sparse PCA for biological integration and interpretable brain network discovery—showcasing NSA-Flow's versatility across ML domains.

4. An [open-source implementation](https://github.com/stnava/nsa-flow) in `pytorch` facilitating easy integration into existing workflows for researchers and practitioners. NSA-flow is also `pip` installable via the `nsa-flow` package and wrapped in `R` via `ANTsR`.

The paper is organized as follows: Section 2 derives the formulation and algorithm; Section 3 details the experimental results; Section 4 discusses limitations and future work; Section 5 gives an overview of software resources.


# Methods

We consider the problem of finding a matrix \( Y \in \mathbb{R}^{p \times k} \) that optimally approximates a target matrix \( X_0 \in \mathbb{R}^{p \times k} \) while satisfying column orthogonality and, optionally, non-negativity. This general formulation is central to a wide range of problems in machine learning and signal processing, including orthogonal dictionary learning, Independent Component Analysis (ICA), and the orthogonal Procrustes problem [@lee2001algorithms; @hyvarinen2000independent; @schonemann1966generalized]. The NSA-Flow optimization problem is defined by the minimization of a composite energy function \( E(Y) \):

\[
\min_{Y \in \mathbb{R}^{p \times k}, Y \ge 0}
E(Y)
= (1 - w) \, L_{fid}(Y, X_0)
+ w \, L_{orth}(Y),
\label{eq:objective}
\]

where \( w \in [0, 1] \) balances the fidelity loss \(L_{fid}\) against the orthogonality loss \(L_{orth}\). Both losses are dimensionless, so no re-weighting, warm-up calibration or data-dependent scaling constant is needed and \(w\) is a convex weight. The fidelity term is the relative squared Frobenius distance
\[
L_{fid}(Y, X_0) = \frac{\| Y - X_0 \|_F^2}{\| X_0 \|_F^2},
\]
which is \(0\) at \(Y = X_0\) and exactly \(1\) at \(Y = 0\).

For the orthogonality loss, let \( G = Y^\top Y / \operatorname{tr}(Y^\top Y) \) be the trace-normalised Gram matrix. We use one of two penalties:
\[
D(Y) = \Big\| G - \tfrac{1}{k} I_k \Big\|_F^2,
\qquad
C(Y) = \frac{1}{k(k-1)} \big\| \widehat{G} - I_k \big\|_F^2,
\qquad
\widehat{G}_{ij} = \langle \hat y_i, \hat y_j \rangle,
\quad
\hat y_i = \frac{y_i}{\max(\|y_i\|, \varepsilon)} .
\]

The identity in \(C\) is not cosmetic and dropping it breaks the functional, so we
write \(C\) out. For any column with \( \|y_i\| \ge \varepsilon \) we have
\( \widehat{G}_{ii} = 1 \) and \( \widehat{G}_{ij} = \cos(y_i, y_j) \), giving
\[
C(Y) = \frac{1}{k(k-1)} \Bigg[ \underbrace{\sum_{i \neq j} \cos^2(y_i, y_j)}_{\text{pairwise angles}}
 + \underbrace{\sum_i \Big( 1 - \min\big(1, \|y_i\|^2/\varepsilon^2\big) \Big)^2}_{\text{floored columns}} \Bigg] ,
\]
where \( \cos(y_i, y_j) = \langle y_i, y_j\rangle / (\|y_i\|\,\|y_j\|) \). The second
sum is *exactly* zero on any \(Y\) whose columns all have norm at least
\(\varepsilon\), which is why the pairwise sum alone is the whole story in the
generic case and why it is tempting to define \(C\) that way. It is not: each
column whose norm has collapsed is charged \( 1/(k(k-1)) \), so a \(Y\) with one
live column and \( k-1 \) dead ones scores
\[
C = \frac{k-1}{k(k-1)} = \frac{1}{k},
\]
\(0.2\) at \(k = 5\) (row four of Table \@ref(tab:props)), whereas the pairwise sum
alone scores it \(0\) --- a dead column is orthogonal to everything. Defining \(C\)
by the pairwise sum therefore reproduces the exact failure that rules out the
off-diagonal defect (§\@ref(percolumn) and row four of Table \@ref(tab:props)),
and any implementation written from that equation inherits it. In the code the
two forms are `angle_defect(V, diagonal=True)` and `diagonal=False`. The variant
without the identity, \( \|\widehat{G} - \operatorname{diag}(\widehat{G})\|_F^2 \),
is used in exactly one place, the signed lifting (`nsa_flow.signed`), where an
empty lobe is the correct answer for a one-signed component and the
reconstruction term is what keeps the basis non-degenerate; everywhere in this
paper \(C\) means the form with the identity.

\(D\) is scale invariant, lies in \([0, 1 - 1/k]\), and vanishes exactly on the scaled Stiefel manifold \( \mathbb{R}_{>0}\cdot St(p,k) \); we report \( \tilde D = D / (1 - 1/k) \) so that it lies in \([0,1]\) and equals \(1\) at rank one. It admits the spectral reading
\[
D = k \operatorname{Var}(\lambda_i(G)) = \frac{1}{\mathrm{EffRank}(Y)} - \frac{1}{k},
\]
so it is a participation-ratio deficit, and \( D \ge 1/r - 1/k \) whenever \(\operatorname{rank}(Y) = r < k\): rank collapse is penalised rather than rewarded.

\(D\) decomposes as
\[
D = \underbrace{\sum_{i \neq j} G_{ij}^2}_{\text{decorrelation}}
  + \underbrace{\sum_i \big(G_{ii} - \tfrac1k\big)^2}_{\text{norm balance}},
\]
so \(D = 0\) requires mutual orthogonality and equal column norms, which is orthonormality. \(C\) requires orthogonality at any column norms. Use \(D\) when the components should carry equal weight and \(C\) when they should not: a disjoint non-negative basis whose columns differ in magnitude scores \(D = 0.14\) and \(C = 2\times 10^{-33}\) (Table \@ref(tab:props)). When the basis is fitted to data the two behave the same, because the reconstruction term already equalises the column norms.

```{r props, results='asis'}
ktab(py_to_r_df(sup$properties_table()) |>
       dplyr::rename(`case` = case, `D (orthonormality)` = D,
                     `off-diagonal only` = d_v1, `C (orthogonality)` = C),
     digits = 4,
     caption = "The three candidate functionals on the structures that separate them. Zero means the functional recognises the structure as what it is looking for. Rows two and three are what this method exists to produce and only the last two columns score them at zero; row four is rank collapse and only the off-diagonal defect rewards it.")
```


The Euclidean gradient of the \(\tilde D\) form is available in closed form,
\[
\nabla_Y E(Y) = (1-w)\,\frac{2 (Y - X_0)}{\|X_0\|_F^2}
 + \frac{w}{1 - 1/k}\cdot\frac{4}{t^2}\Big[ Y S - \frac{\|S\|_F^2}{t} Y \Big],
\qquad S = Y^\top Y,\; t = \operatorname{tr} S,
\]
requiring no factorisation. It satisfies the Euler identity \( \langle \nabla D, Y\rangle = 0 \), since \(D\) is homogeneous of degree zero, so the orthogonality term cannot alter \( \|Y\|_F \); and \( \|\nabla D\|_F \le 8 / \|Y\|_F \). For \(C\) the gradient is tangential to each column's sphere separately, so \(C\) cannot alter *any* individual column norm.

## The Optimization Challenge and the NSA-Flow Approach

A plain Euclidean gradient step, \( Y \leftarrow Y - \eta \nabla_Y E(Y) \), does not respect the constraints. The conventional remedy is Riemannian optimisation: project the gradient onto the tangent space of the Stiefel manifold and apply a **retraction** to return to it [@absil2008optimization; @boumal2023intro]. Retractions are theoretically sound but require an SVD or an inverse matrix square root at every step, and enforcing exact feasibility can fight the fidelity term by abruptly correcting the iterate's geometry.

NSA-Flow keeps non-negativity as a hard constraint, which its exact projection makes cheap, and treats orthogonality as an objective. Nothing is retracted and no factorisation appears in the loop. Figure \@ref(fig:pollen) shows how the iterates evolve with \(w\).

### Solver: spectral projected gradient

The feasible set \( \{ Y \ge 0 \} \) is closed and convex and its Euclidean projection is the elementwise clamp \( \Pi(Y) = \max(0, Y) \), so the problem is a smooth objective on a convex set and admits a projected-gradient method. We use the spectral projected gradient of Birgin, Martínez and Raydan [@birgin2000nonmonotone]: Barzilai--Borwein step lengths [@barzilai1988two]
\[
t_n = \frac{\langle s_n, s_n\rangle}{\langle s_n, r_n\rangle},
\qquad s_n = Y_n - Y_{n-1},\quad r_n = \nabla E(Y_n) - \nabla E(Y_{n-1}),
\]
safeguarded by an Armijo backtracking line search on the projected step
\[
Y_{n+1} = \Pi\big(Y_n - t_n \nabla E(Y_n)\big),
\qquad
E(Y_{n+1}) \le E(Y_n) - \sigma \frac{\|Y_{n+1} - Y_n\|_F^2}{t_n}.
\]
The line search makes the energy monotonically non-increasing by construction, and for a continuously differentiable \(E\) on a closed convex set the method converges to a stationary point.

### Stationarity certificate

The quantity
\[
\gamma_n = \frac{\|\Pi(Y_n - t_n \nabla E(Y_n)) - Y_n\|_F}{t_n}
\]
is the norm of the projected-gradient mapping. It vanishes at a stationary point of the constrained problem and remains valid on the boundary, where non-negative solutions lie and where the gradient norm does not vanish. The solver reports \(\gamma_n\) and which stopping rule fired: certificate threshold, line-search exhaustion or iteration cap. Conditioning degrades as \( w \to 1 \), where the problem approaches its combinatorial limit: convergence takes tens of iterations at \(w = 0.5\), hundreds at \(w = 0.9\), and thousands beyond \(w = 0.99\).

At \( w = 1 \) the fidelity term vanishes and \(D\) is scale invariant, so the scale of \(Y\) is unconstrained: every scaled Stiefel matrix is optimal and nothing selects among clusterings. The implementation warns at \(w = 1\) and reports the observed scale drift.

### Why the penalty must be normalised per column, not globally {#percolumn}

The unnormalised penalty \( \| Y^\top Y - I \|_F^2 \) is sensitive to the norm of \(Y\): scaling \( Y \to cY \) scales it by \(c^4\), which makes the optimisation poorly conditioned and the step size hard to choose. An earlier version of this work therefore used a globally normalised off-diagonal defect,
\[
L_{orth, inv}(Y) = \frac{\| Y^\top Y - \operatorname{diag}(\operatorname{diag}(Y^\top Y)) \|_F^2}{\|Y\|_F^4},
\]
described as measuring the cosines of the angles between columns. It does not. Dividing by the global \( \|Y\|_F^4 \) rather than by \( \|y_i\|^2\|y_j\|^2 \) gives a quantity invariant to global rescaling but not to rescaling columns individually, so it is not a function of the angles. Its zero set is also too large. A matrix with one non-zero column and \(k-1\) zero columns attains \( L_{orth, inv} = 0 \) exactly, since a zero column is orthogonal to everything, so the functional is minimised by rank collapse. \(C\) scores that matrix \(0.2\) and \(D\) scores it \(0.8\).

\(C\) normalises each column, so it is a function of the angles and invariant to per-column rescaling, which is the gauge freedom of a basis; its diagonal charges a column whose norm has collapsed. \(D\) adds the norm-balance term and its floor \(D \ge 1/r - 1/k\) at rank \(r\) penalises collapse. The off-diagonal defect alone does neither and is not used here.

### Relationship to Alternative Manifold Optimization Methods

*   **Cayley Transform:** An alternative for preserving orthogonality is the **Cayley transform**, which defines an exact retraction-free update. For a skew-symmetric matrix \( A = \text{grad}E(Y) Y^\top - Y (\text{grad}E(Y))^\top \), the update \(Y^{(t+1)} = ( I - \frac{\eta}{2} A )^{-1} ( I + \frac{\eta}{2} A ) Y^{(t)}\) exactly preserves orthogonality [@gao2019parallelizing]. However, it requires solving a \(p \times p\) linear system, making it computationally prohibitive for large \(p\).

*   **Riemannian Optimization Frameworks:** Standard toolboxes like `Manopt` [@boumal2014manopt] implement sophisticated algorithms like Riemannian trust-region and conjugate gradient methods. NSA-Flow instead avoids the manifold machinery altogether --- orthogonality is an objective, not a constraint --- which makes it a drop-in component for existing ML or deep-learning pipelines and lets the degree of constraint satisfaction be chosen rather than fixed.

### Computational Complexity

```{r complexitytable, echo=FALSE}
cx <- data.frame(
  Method = c("Euclidean gradient descent", "Full polar retraction",
             "NSA-Flow", "Cayley transform"),
  `Dominant operation` = c("gradient",
                           "gradient + SVD or polar factor",
                           "gradient + clamp, no factorisation",
                           "linear solve, $p \\times p$"),
  `Cost per iteration` = c("$O(pk^2)$", "$O(pk^2)$",
                           "$O(pk^2)$, or $O(npk)$ matrix-free",
                           "$O(p^3)$"),
  Orthogonality = c("not enforced", "exact", "set by $w$", "exact"),
  check.names = FALSE)
ktab(cx, caption = "Cost and orthogonality guarantee for NSA-Flow and related updates. The asymptotic cost of NSA-Flow matches a retraction method's, but the per-step work is a few matrix products rather than a factorisation.",
     escape = FALSE, align = c("l", "l", "l", "l"))
```


For tall-skinny matrices (\( p \gg k \)) each iteration costs one Gram product and two \([p,k]\times[k,k]\) products, i.e.\ \( O(pk^2) \), with no SVD, eigendecomposition or QR anywhere in the loop --- the asymptotic cost is the same as a retraction method's, but the constant is much smaller and the per-step work is a few matrix multiplications rather than a factorisation. When the basis is fitted directly to data with \( p > n \) the Gram matrix is never formed either: every quantity follows from \( XV \) and \( X^\top(XV) \), giving \( O(npk) \) per iteration and removing an allocation that is \(0.4\) GB at \( p = 7129 \) and \(3.2\) GB at \( p = 20000 \).  One parameter sets orthogonality, and with it sparseness, as the Results section shows.


```{r pollen, echo = FALSE, fig.cap = "Illustration of the NSA-Flow optimization as a function of $\\omega$.  The colored manifold is a conceptual representation of the Stiefel manifold, with the curves representing optimization paths for evolving $Y$.  When $\\omega$ is small the orthogonality term is weakly weighted and the iterate is free to deviate from orthonormality; when $\\omega$ is large the term dominates and the solution is drawn toward the manifold."}
knitr::include_graphics("figs/nsa_flow_manifolds_fig.pdf")
```

### Implementation

NSA-Flow is implemented in `pytorch`; the results here call it directly. The library is five modules and under a thousand lines, down from eight modules and about 2400.

The refinement entry point takes a target \( X_0 \), an optional initialisation and the weight \( w \). The estimation entry point takes the data matrix \( X \) and the number of components \( k \). Gone are the retraction type, optimizer backend, learning-rate strategy, aggression and warm-up parameters, and the separate fidelity and orthogonality switches. These compensated for an energy whose two terms were not comparably scaled and are unnecessary once both are dimensionless.

Each iteration computes a fused value and gradient sharing one Gram product, takes a Barzilai--Borwein step and accepts it through an Armijo line search on the projected step, so the energy is monotone. Non-negativity is applied by exact projection. Per-iteration diagnostics are the energy, the two loss terms, the step length and the gradient-mapping norm. The returned object carries the stopping reason, the certificate, the iteration count, the effective rank and the scale ratio.

Two numerical details. The orthogonality defect is computed as \( \|G - I/k\|_F^2 \) directly rather than as \( \|G\|_F^2 - 1/k \); the latter is algebraically identical but suffers catastrophic cancellation near the optimum and returns small negative values. And when a polar factor must be differentiated, the derivative is obtained by solving the Sylvester equation, whose denominators are \( h_i + h_j > 0 \), rather than through the SVD backward pass, whose denominators are \( \sigma_i^2 - \sigma_j^2 \) and which returns `NaN` for *every* entry when singular values repeat --- as they do for any orthogonal initialisation.

The test suite asserts the propositions of this section: the zero set, the bounds, the spectral identity, the rank-collapse floor, the invariances, the disjoint-support equivalence, agreement of every closed-form gradient with automatic differentiation, and equality of the Gram and matrix-free paths. Centre and scale the input and choose \( w \) for the application. There is nothing else to set.

```{r nsaflowchart, fig.cap="NSA-Flow Algorithm Workflow", fig.width=6, fig.height=4,echo=FALSE}
nsa_flow_flowchart <- function(
  node_fill_color = "lightblue",
  node_font_color = "black",
  edge_color = "navy",
  font_name = "Helvetica",
  node_shape = "rectangle",
  graph_rankdir = "TB",
  fontsize = 10
) {
  library(DiagrammeR)
  graph_spec <- paste0("
digraph nsa_flow_algorithm {
  graph [rankdir = ", graph_rankdir, ", fontsize = ", fontsize, ", splines = ortho]
  node [shape = ", node_shape, ", style = filled, fillcolor = ", node_fill_color, ", 
        fontcolor = ", node_font_color, ", fontname = ", font_name, ", fontsize = ", fontsize, "]
  edge [color = ", edge_color, ", fontsize = ", fontsize, "]
  A [label = 'Initialize Y_0\\n(Random or SVD-based)']
  B [label = 'Set w and tolerance\\n(no learning rate, no optimizer choice)']
  C [label = 'Fused value and gradient\\n∇E = (1-w)∇L_fid + w∇L_orth\\n(one Gram product, no factorisation)']
  D [label = 'Barzilai-Borwein step length\\nt = <s,s> / <s,r>']
  E [label = 'Projected trial step\\nY+ = max(0, Y - t ∇E)']
  F [label = 'Armijo test\\nE(Y+) <= E(Y) - σ||Y+-Y||²/t ?']
  G [label = 'Halve t\\nand retry', fillcolor = lightyellow]
  I [label = 'Stationarity certificate\\nγ = ||Y+ - Y|| / t']
  J [label = 'Accept Y+\\nRecord diagnostics\\n(iter, energy, fid, defect, γ)']
  K [label = 'Output Y and stop_reason\\n(grad_map | line_search | max_iter)']
  A -> B [label = 'Setup']
  B -> C [label = 'Start iteration']
  C -> D [label = 'Step length']
  D -> E [label = 'Trial']
  E -> F [label = 'Test']
  F -> G [label = 'Reject', style = 'dashed']
  G -> E [label = 'Retry', style = 'dashed']
  F -> I [label = 'Accept']
  I -> J [label = 'γ > tol', style = 'dashed']
  J -> C [label = 'Next iteration']
  I -> K [label = 'γ <= tol']
}
")
  grViz(graph_spec)
}
nsa_flow_flowchart(node_fill_color = "lightblue", edge_color = "navy", fontsize = 10)



#' Compute Sparsity Level of a Matrix
#'
#' @param M A numeric matrix.
#' @param tol Numeric tolerance: entries with absolute value below `tol`
#'        are treated as zero (default = 1e-8).
#' @return A numeric value between 0 and 1, representing the fraction of
#'         (near-)zero entries. Higher = more sparse.
#' @examples
#' M <- matrix(c(1, 0, 0, 2, 0, 0), nrow = 2)
#' sparsity_level(M)
#' # [1] 0.6666667
sparsity_level <- function(M, tol = 1e-8) {
  stopifnot(is.matrix(M), is.numeric(M))
  total <- length(M)
  zeros <- sum(abs(M) < tol)
  sparsity <- zeros / total
  return(sparsity)
}

```

## Sparse PCA via NSA-Flow

Sparse principal component analysis (SPCA) finds a sparse basis \( Y \in \mathbb{R}^{p \times k} \) that maximizes the variance explained in a data matrix \( X \in \mathbb{R}^{n \times p} \). We can formulate SPCA as a regularized optimization over $Y$ :

\[
\min_{Y} \left\{ -\frac{1}{2n} \text{tr}(Y^\top X_c^\top X_c Y) + \lambda R(Y) \right\},
\]
where \( X_c \) is the data matrix, \( \lambda \ge 0 \) is a regularization parameter controlling the regularization penalty, and \( R(Y) \) is a suitable regularization function.  The smooth component of the objective is the negative explained variance, \( f(Y) = -\frac{1}{2n} \text{tr}(Y^\top S Y) \), where \( S = X_c^\top X_c \) is the covariance matrix. The Euclidean gradient of this term is:
\[
\nabla_Y f(Y) = -\frac{1}{n} S Y.
\]


We optimize this energy in two main steps:

**Step 1: Gradient Descent with Line Search**
A candidate update \( Z^{(t)} \) is computed by taking a step along the negative gradient direction from the current iterate \( Y^{(t)} \):
\[
Z^{(t)} = Y^{(t)} - \alpha^{(t)} \nabla_Y f(Y^{(t)}),
\]
where the step size \( \alpha^{(t)} \) is determined by an Armijo-type backtracking line search to ensure sufficient decrease in the objective function.

**Step 2: Proximal Step and Orthogonality Enforcement**
The non-smooth regularization term arising from $R$ is handled in a proximal step. The method supports two types of proximal updates:

*   **`proximal_type = "basic"` (Proximal Thresholding):** A standard soft-thresholding operator is applied to the candidate \( Z^{(t)} \) to induce sparsity, followed by an optional non-negativity projection:
    \[
    Y^{(t+1)} = \text{prox}_{\alpha\lambda, \|\cdot\|_1}(Z^{(t)}) = \text{sign}(Z^{(t)}) \odot \max(|Z^{(t)}| - \alpha^{(t)}\lambda, 0).
    \]
    This standard approach decouples the sparsity and orthogonality steps. We  re-orthogonalize via QR at the beginning of each iteration when this algorithmic path is taken.

*   **`proximal_type = "nsa_flow"` (Proximal Flow):** A more sophisticated proximal step is performed by invoking the NSA-Flow algorithm. The candidate matrix \( Z^{(t)} \) serves as the target for an inner NSA-Flow optimization loop:
    \[
    Y^{(t+1)} = \arg\min_{U \ge 0} \left\{ \frac{1}{2} \| U - Z^{(t)} \|_F^2 (1-w) + \, \text{Orth}(U) w \right\}.
    \]
    This subproblem simultaneously encourages fidelity to the gradient-updated iterate \( Z^{(t)} \), promotes sparsity through non-negativity, and enforces column orthogonality via the flow. This approach integrates the constraints more tightly into the optimization, providing a unified update that respects both the geometry and the regularization.

The algorithm terminates when the projected-gradient mapping norm falls below the tolerance, and reports which stopping rule fired. There is no learning-rate schedule: the Barzilai--Borwein step is scale-free and the Armijo line search halves it until the energy decreases, so the energy is monotone without a patience parameter. The final output is the set of sparse loadings \( Y \) that achieved the lowest energy during the optimization.  This is one example where NSA-Flow can be integrated as a proximal operator within a broader optimization framework to enforce orthogonality and non-negativity.  Other approaches (shown below) more directly use NSA-Flow as a standalone method for matrix approximation; for example, by directly approximating the loading matrix of PCA with the bases derived from NSA-Flow.  Both SPCA and the latter approach are demonstrated in the Results section.  The solver used here is the Python package; `ANTsR`'s `nsa_flow_pca` and `nsa_flow_pca_fa` wrap the superseded version and are not what produced these results.


# Results

## Choice of downstream model in the evaluations {#evaluation}

The comparisons below use random forests. The reason is an identity. Fit a linear
model to scores \( XV \). For any invertible \( M \),
\[
\operatorname{span}(XVM) = \operatorname{span}(XV),
\]
so the fitted values are unchanged when \( V \) is replaced by \( VM \): the
coefficients absorb \( M^{-1} \). Ordinary least squares, ridge and logistic
regression are all invariant to reparametrisation of the basis. They see the
subspace and not the axes.

NSA-Flow refines a basis and stays near it, so \( V_{\text{NSA}} \) and
\( V_{\text{PCA}} \) span almost the same subspace. A linear comparison of the two
will tie whatever the axes look like. We measured this on the ADNI thickness
data: replacing \( V \) by \( VM \) for random invertible \( M \) changes
cross-validated linear \( R^2 \) by \( 0.0000 \) and random-forest \( R^2 \) by
\( 0.057 \). A tree splits on single coordinates, so its fit depends on the axes.

Sensitivity is not the same as preference, and the direction is data dependent.
On ADNI thickness with cognitive scores as outcomes, the forest prefers the
NSA-Flow axes: out-of-sample \( \Delta R^2 \) over an age, sex, education and
APOE4 baseline is \( 0.043 \) higher than PCA, on \( 7 \) to \( 9 \) of \( 9 \)
outcomes across three seeds. On ADNI regional volumes, `digits` and
`breast_cancer` with binary labels, the forest prefers the PCA axes by about
\( 0.010 \) AUC. Under logistic regression all of these differences shrink toward
zero, as the identity requires.

We report linear results where the subspace is the question and tree results
where the axes are. A tie under a linear model is not evidence about a basis.

## Toy Example: Decomposing a Small Mixed-Signal Matrix

To intuitively illustrate NSA-Flow, consider a toy 4x3 matrix \( X_0 \) representing mixed signals: each column is a nonnegative orthogonal basis vector (e.g., distinct patterns), but observed with noise and scaling. NSA-Flow approximates an orthogonal nonnegative basis \( Y \) close to \( X_0 \).  Figure \@ref(fig:toyexample) shows the results of applying NSA-Flow to this toy example, starting from a random initialization \( Y_0 \).

```{r toyexample, echo=FALSE, fig.width=6, fig.height=3,fig.cap="NSA-Flow applied to a toy 4x3 matrix with noisy orthogonal nonnegative patterns."}
set.seed(42)
# True orthogonal nonnegative basis
true_Y <- matrix(c(1,0,0,0, 0,1,0,0, 0,0,1,0), 4, 3)  # Nearly orthogonal, nonnegative
true_Y <- true_Y + matrix(runif(12, 0, 0.1), 4, 3)  # Add small perturbations
true_Y <- pmax(true_Y, 0)
true_Y <- true_Y %*% solve(chol(crossprod(true_Y)))  # Orthogonalize

# Noisy target
X0_toy <- true_Y + matrix(rnorm(12, 0, 0.2), 4, 3)
X0_toy <- pmax(X0_toy, 0)  # Ensure nonnegative

# Initial random guess
Y0_toy <- matrix(runif(12, 0, 1), 4, 3)

# Apply NSA-Flow with balanced weights
# X0_toy=X0_toy/norm(X0_toy, "F")*0.1  # Normalize
omega_default = 0.5
# if ( ! exists("ini_default") ) 
## The v1 menus -- learning-rate strategy, optimizer backend, retraction type,
## aggression, warm-up, separate fidelity/orthogonality type switches -- have no
## counterpart in v2 and are gone.  What remains is w, the iteration cap and the
## tolerance.
nsa_default <- function(Y0, w = omega_default, X0 = NULL, verbose = FALSE,
                        fidelity = "auto") {
  nsa_flow_autograd(
    Y0 = Y0,
    X0 = X0,
    w = w,
    max_iter = 2000,
    tol = 1e-9,
    apply_nonneg = TRUE,
    verbose = verbose,
    fidelity = fidelity
  )
}
#
res_toy <- nsa_default(Y0 = X0_toy, X0 = true_Y, w = omega_default )
# Visualize

library(ggplot2)
library(reshape2)
library(patchwork)   # or cowplot

# Helper: convert a matrix to a ggplot heatmap
make_heatmap <- function(mat, title) {
  df <- melt(mat)
  ggplot(df, aes(Var2, Var1, fill = value)) +
    geom_tile() +
    scale_fill_viridis_c(option = "plasma") +
    coord_equal() +
    theme_minimal(base_size = 8) +
    theme(
      axis.title = element_blank(),
      axis.text = element_blank(),
      panel.grid = element_blank(),
      plot.title = element_text(face = "bold", hjust = 0.5)
    ) + theme(legend.position = "none") +
    ggtitle(title)
}

# Create the four ggplots
p1 <- make_heatmap(X0_toy,  "Noisy Target X0")
p2 <- make_heatmap(Y0_toy,  "Initial Y0")
p3 <- make_heatmap(res_toy$Y, "NSA-Flow Y (w=0.5)")
p4 <- make_heatmap(true_Y,  "True Basis")

# Arrange them in a 1x4 grid for better visibility
(p1 | p2 | p3 | p4 ) 
# Metrics
toy_metrics <- data.frame(
  Metric = c("Recon Error", "Orth Residual", "Neg Violation"),
  Value = c(frob(res_toy$Y - X0_toy), orth_residual(res_toy$Y), neg_violation(res_toy$Y))
)
# kable(toy_metrics, digits = 4, caption = "Toy Example Metrics")
```

**Interpretation**: Starting from a random \( Y_0 \), NSA-Flow recovers a basis close to the true orthogonal nonnegative patterns in \( X_0 \), with low reconstruction error and near-zero orthogonality/nonnegativity residuals. This captures the essence: extracting interpretable, disjoint components from noisy data. 





## Behaviour of the orthogonality weight

The earlier version of this work compared optimizer backends (ASGD, Adam, AdaGrad and others) and swept an interpolation parameter between a Euclidean step and a polar retraction across tall, wide and square matrices. Neither choice exists now: there is one solver and nothing is retracted. The two figures reporting those comparisons have been removed rather than recaptioned, since they were produced by code that is gone. What remains to report is how \( w \) behaves, which is Section \@ref(sparsity) and the convergence traces in Figure \@ref(fig:stiefelsweepfull).

The comparison against the retraction scheme was rerun rather than dropped. On the earlier study's synthetic problem (\( p = 200 \), \( k = 8 \)) and scored on its own metrics, the present solver attains an orthogonality residual of \( 0.011 \) at \( w = 0.999 \) against \( 0.016 \) for polar and \( 0.019 \) for soft retraction, and \( 0.0018 \) at \( w = 0.9999 \); it matches the best retraction on that study's energy (\( 5.899 \) against \( 5.845 \), with projected gradient at \( 59.5 \)); it returns the same basis from twelve random restarts where polar retraction has a spread of \( 9 \times 10^{-4} \); and its runtime is nearly flat in \( p \) since no factorisation is performed. The orthogonality residual is set by \( w \) rather than fixed by the operator.

## Sparsity as a Function of Orthogonality via Weight Parameter \( w \) {#sparsity}

Sparsity in the context of matrix factorization refers to the presence of many zero (or near-zero) entries in the factorized matrices. In NSA-Flow, sparsity is not directly enforced through explicit penalties (like L1 regularization) but emerges as a consequence of promoting orthogonality among the columns of the matrix \( Y \).  The parameter $\omega$ serves as a trade-off weight between data fidelity and orthogonality regularization. Orthogonality is measured at the whole-matrix level where lower values indicate closer alignment to an orthogonal (or near-orthogonal) structure.

By adjusting $\omega$, sparsity is indirectly controlled through this global orthogonality constraint:

- **Low \( w \) (e.g., 0.05–0.25)**: Prioritizes fidelity to the input data, resulting in denser matrices with higher entry correlations across columns. Sparsity remains low, as the optimization allows overlapping patterns to preserve original structure, leading to "blurry" approximations.

- **Increasing \( w \) (e.g., 0.5–0.75)**: Strengthens orthogonality enforcement, promoting decorrelated columns. This induces sparsity by concentrating non-zero entries into disjoint patterns, reducing overlap and yielding moderate sparsity.

- **High \( w \) (e.g., 0.95)**: Dominates with orthogonality, forcing near-orthogonal columns that are highly sparse (\( \approx 0.9 \)) and crisp, but potentially over-constrained, risking loss of fidelity to the original data.

Matrix-level orthogonality produces sparsity without per-entry penalties. In synthetic experiments, heatmaps of the optimised matrices transition from diffuse (low \( w \)) to sharp and disjoint (high \( w \)) (Figure \@ref(fig:stiefelsweepfull)). The convergence traces in Figure \@ref(fig:convergetraces) show stable optimisation across \( w \) values, confirming the parameter's role in balancing these objectives.  Note that the exact sparsity levels depend on data characteristics and initialization, but the trend of increasing sparsity with higher \( w \) is consistent regardless of whether data is thin, wide, or square.

```{r synthsetup,fig.height=8,fig.width=5.3, echo=FALSE, message=FALSE}
generate_synth_data <- function(p=40, k=3, corrval=0.3,noise=0.5, sparse_prob=0.2, include_neg=FALSE) {
  V0 <- qr.Q(qr(matrix(rnorm(p*k), p, k)))
  Y0 <- V0 + noise * matrix(rnorm(p*k), p, k)
  Y0 <- matrix(rnorm(p*k), p, k)+1
  library(MASS)
  # Target covariance
  Sigma <- matrix(corrval, k, k)
  diag(Sigma) <- 1  # variances = 1
  # Generate correlated datap
  Y0 <- mvrnorm(n = p, mu = rep(0, k), Sigma = Sigma)
  if (include_neg) Y0 <- Y0 - 0.5
  if (sparse_prob > 0) Y0[runif(p*k) < sparse_prob] <- 0
  X0 <- pmax(Y0 + 0.2 * matrix(rnorm(p*k), p, k), 0)
  list(Y0=Y0, X0=X0, V0=V0)
}

set.seed(123)
p <- 200; k <- 40
X0 = generate_synth_data( p, k, corrval=0.35, noise=0.05, sparse_prob=0.0, include_neg=FALSE )$Y0
####
```


```{r stiefelsweepfull, fig.width=6.5,  echo=FALSE, message=FALSE,fig.cap="A synthetic dataset is generated with controlled correlation and noise levels to evaluate NSA-Flow's performance across different orthogonality weights ($\\omega$). The data matrix $X_0$ is approximated with NSA-Flow to reveal underlying orthogonal structures, allowing assessment of how varying $\\omega$ influences the sparsity and orthogonality of the resulting factorization."}
###
w_seq <- c( 0.005,  0.05, 0.1, 0.2, 0.5 )
w_seq <- c( 0.001,  0.005, 0.01, 0.05, 0.25 )
w_seq <- c( 0.1,  0.25, 0.5, 0.75, 0.9 )
mytit = paste0("w = ", round(w_seq,3))
mats <- list()
convergeplots <- list()
for(i in seq_along(w_seq)) {
  w_val <- w_seq[i]
  res_soft_w <- nsa_default( X0, w = w_val, verbose = FALSE )
  mytit[i] <- paste0("w = ", round(w_val, 3), ', orth = ', 
    round(invariant_orthogonality_defect(res_soft_w$Y),4), ', w.spar = ',
    1.0-round(sum(res_soft_w$Y/max(res_soft_w$Y) > quantile(res_soft_w$Y,0.1))/length(res_soft_w$Y),3))
  mats[[i]] <- res_soft_w$Y
  convergeplots[[i]] <- res_soft_w$plot+labs(title =mytit[i], subtitle = NULL)+
  theme_minimal(base_size = 8)+theme(legend.position = "top")
}
allvals <- unlist(lapply(mats, function(m) as.numeric(m)))
rng <- c(   
    quantile(c(as.numeric(X0), allvals), c(0.05), na.rm = TRUE) , 
    quantile(c(as.numeric(X0), allvals), c(0.95), na.rm = TRUE))
cols <- rev(colorRampPalette(brewer.pal(9,"YlGnBu"))(120))
swplots <- vector("list", length(mats)+1)
rows_show <- 1:nrow(X0)
ogtit = paste0("Original ",', orth = ', 
    round(invariant_orthogonality_defect(X0),4), ', w.spar = ',
    1.0-round(sum(X0/max(X0) > quantile(X0,0.1))/length(X0),3))
swplots[[1]] <- pheatmap(X0[rows_show,], cluster_rows = FALSE, cluster_cols = FALSE, color = cols, main = ogtit, fontsize=5, silent = TRUE, breaks = seq(rng[1], rng[2], length.out = 121))
for(i in seq_along(mats)) {
  swplots[[i+1]] <- pheatmap(mats[[i]][rows_show,], cluster_rows = FALSE, cluster_cols = FALSE, color = cols, main = mytit[i], fontsize=5, silent = TRUE, breaks = seq(rng[1], rng[2], length.out = 121))
}
grid.arrange(grobs = lapply(swplots, function(x) x$gtable), ncol = 3)
```

```{r convergetraces, fig.width=6.5, fig.height=4.5, echo=FALSE, message=FALSE, fig.cap="Energy terms against iteration at four values of the orthogonality weight. Fidelity and the orthogonality defect are shown on a log scale; the solver is monotone in the total energy by construction, so both curves are traced from the same accepted steps."}
## Emitted as its own chunk rather than a second plot in the sweep chunk: two
## figures from one chunk share a caption and receive no cross-reference label,
## which is why a reference to them rendered as "??".
if (length(convergeplots) >= 4) {
  grid.arrange(grobs = convergeplots[c(1, 2, 3, 5)], ncol = 2)
} else {
  message("convergence traces unavailable: ", length(convergeplots), " plots")
}
```


<!--
## Illustration of Weight Parameters

```{r wfidsweep,cache=FALSE}
set.seed(1)
p_test <- 100; k_test <- 20
Y0_test <- matrix(rnorm(p_test * k_test), p_test, k_test)
Y0_test <- Y0_test / frob(Y0_test)
ws <- c(0:50)/200
ws[1]=1e-5
results <- lapply(ws, function(w) {
  out <- nsa_default(Y0_test, w = w )
  data.frame(w = w, frob_error = frob(out$Y - Y0_test), orth_error = orth_residual(out$Y))
})
df <- do.call(rbind, results)
# kable(df, digits = 6, caption = "Table 5: Effect of w on fidelity and orthogonality errors")
#
#
# Load libraries
library(ggplot2)
library(dplyr)
library(scales)

# Normalize orth_error for dual-axis plotting
scale_factor <- max(df$frob_error) / max(df$orth_error)
df <- df %>% mutate(orth_scaled = orth_error * scale_factor)

# Plot
ggplot(df, aes(x = w)) +
  geom_line(aes(y = frob_error, color = "Frobenius Error"), linewidth = 1.3) +
  geom_point(aes(y = frob_error, color = "Frobenius Error"), size = 2.5) +
  geom_line(aes(y = orth_scaled, color = "Orthogonality Error"), linewidth = 1.3, linetype = "dashed") +
  geom_point(aes(y = orth_scaled, color = "Orthogonality Error"), size = 2.5, shape = 17) +
  scale_y_continuous(
    name = "Frobenius Error (lower is better)",
    sec.axis = sec_axis(~ . / scale_factor,
                        name = "Orthogonality Error (lower is better)")
  ) +
  scale_color_manual(values = c("Frobenius Error" = "#1f78b4", "Orthogonality Error" = "#e31a1c")) +
  labs(
    title = "Effect of Weight Parameter (w) on Fidelity and Orthogonality",
    subtitle = "Frobenius and orthogonality errors vs. regularization weight w",
    x = "Regularization Weight (w)",
    color = ""
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5, color = "gray40"),
    axis.title.y.left = element_text(color = "#1f78b4", face = "bold"),
    axis.title.y.right = element_text(color = "#e31a1c", face = "bold"),
    axis.text = element_text(color = "gray25"),
    legend.position = "top",
    legend.title = element_blank(),
    legend.text = element_text(size = 12)
  )





```

**Interpretation**: As \( w \) increases (more orthogonality emphasis), Frobenius error increases linearly as fidelity is traded off, but orth error decreases, showing the trade-off. At w=0.09, balanced low errors; useful for tuning based on application needs (e.g., high w for decorrelation-heavy tasks).

-->

## Applications

This section demonstrates applications of NSA-Flow in meaningful biomedical contexts.  We first demonstrate performance of Sparse PCA implemented with NSA-Flow on real data from the Golub leukemia dataset [@golub1999molecular].  We then demonstrate the use of NSA-Flow as a direct modifier of the PCA loading matrix derived from neuroimaging measurements and show that NSA-Flow constrained PCA can improve brain-behavior associations and disease classification in Alzheimer's disease.

```{r subtypesetup, warning=FALSE, message=FALSE,fig.cap="Synthetic gene expression data is generated as mixtures of orthogonal nonnegative gene modules associated with each disease subtype, plus noise. NSA-Flow is applied for dimensionality reduction, and the resulting features are used in a multinomial logistic regression classifier to predict disease subtypes. The performance is compared against NMF, PCA (with absolute values), and raw features using 5-fold cross-validation accuracy. Here, t-SNE is used to visualize the raw data as a 2D projection.", fig.width=5,eval=FALSE}



# Set seed for reproducibility
set.seed(2025)

# Parameters
n_samples <- 50      # Number of samples
n_features <- 10     # Number of features
k_basis <- 5        # Number of basis components
n_classes <- 3       # Number of classes (subtypes)

# True nonnegative basis (features x components)
true_basis <- matrix(0, nrow = n_features, ncol = k_basis)
rows_per_component <- ceiling(n_features / k_basis)  # Approx rows per component
for (i in 1:k_basis) {
  start <- ((i-1) * rows_per_component + 1) %% n_features
  if (start == 0) start <- n_features  # Handle modulo edge case
  end <- min(start + rows_per_component - 1, n_features)
  if (start <= end) {
    true_basis[start:end, i] <- runif(end - start + 1, 0.5, 1)
  } else {
    # Wrap around to beginning if start > end
    true_basis[start:n_features, i] <- runif(n_features - start + 1, 0.5, 1)
    true_basis[1:(end %% n_features), i] <- runif((end %% n_features), 0.5, 1)
  }
}
# Normalize to approximate orthonormal basis while preserving nonnegativity
true_basis <- pmax(true_basis / sqrt(colSums(true_basis^2)), 0)  # Normalize columns to unit length

# Generate subtypes (balanced classes)
subtypes <- rep(1:n_classes, length.out = n_samples)  # Evenly distribute classes
subtypes <- sample(subtypes, n_samples)  # Shuffle to randomize order
labels <- factor(subtypes, labels = c("G0", "G1", "G2"))  # Name classes for clarity

# Mixing coefficients biased by subtype (samples x components)
mix_coeffs <- matrix(runif(n_samples * k_basis, 0, 1), nrow = n_samples, ncol = k_basis)
for (i in 1:n_classes) {
  idx <- subtypes == i
  component_idx <- ((i-1) %% k_basis) + 1
  mix_coeffs[idx, component_idx] <- mix_coeffs[idx, component_idx] + 0.5
}
mix_coeffs <- pmax(mix_coeffs, 0)  # Ensure nonnegativity

# Generate data (samples x features)
data_X <- mix_coeffs %*% t(true_basis) + matrix(rnorm(n_samples * n_features, 0, 0.05), n_samples, n_features)
data_X <- pmax(data_X, 0)  # Ensure nonnegativity

# Verify dimensions and data

# Compute random assignment accuracy (multiclass baseline)
class_props <- prop.table(table(labels))
random_accuracy <- sum(class_props^2)

# t-SNE visualization
tsne_result <- Rtsne(data_X, dims = 2, perplexity = min(30, n_samples/4), verbose = FALSE, max_iter = 500)
tsne_df <- data.frame(
  tSNE1 = tsne_result$Y[, 1],
  tSNE2 = tsne_result$Y[, 2],
  Subtype = labels
)

# Plot t-SNE
p <- ggplot(tsne_df, aes(x = tSNE1, y = tSNE2, color = Subtype)) +
  geom_point(size = 2) +
  scale_color_manual(values = c("G0" = "#4ECDC4", "G1" = "#FF6B6B", "G2" = "#1F77B4")) +
  labs(title = "t-SNE Visualization of Synthetic Data by Subtype",
       x = "t-SNE Dimension 1",
       y = "t-SNE Dimension 2",
       color = "Subtype") +
  theme_minimal( base_size = 8 ) +
  theme(legend.position = "bottom")
print(p)

# NMF
library(NMF)
nmf_fit <- nmf(t(data_X), rank = k_basis, method = "Frobenius")
reduced_nmf <- data_X %*% basis(nmf_fit)

# PCA (abs)
pca_fit <- prcomp(data_X, rank. = k_basis)
reduced_pca <- (predict(pca_fit, data_X))

# Raw
reduced_raw <- data_X

# CV accuracy function
cv_accuracy <- function(features, labels) {
  if ( is.null(colnames(features)) ) colnames(features) <- paste0("V", 1:ncol(features))
  train_control <- trainControl(method = "repeatedcv", number = 5, repeats=10)
  model <- train(features, labels, method = "multinom", trControl = train_control, trace = FALSE)
  data.frame( Accuracy = mean(model$results$Accuracy), AccuracySD = mean(model$results$AccuracySD ))
}


# NSA-Flow with multiple w for sensitivity
acc_pca <- cv_accuracy(reduced_pca, labels)
acc_nmf <- cv_accuracy(reduced_nmf, labels)
acc_raw <- cv_accuracy(reduced_raw, labels)

w_values = c( 0.2, 0.5 )
odf=data.frame()
for ( ww in  w_values ) {
  ## Comparable to the NMF and PCA baselines above: obtain a p x k basis and
  ## project the data onto it.  Previously this refined the n x p data matrix
  ## itself and then ran PCA on the result, which is not a basis at all, has no
  ## out-of-sample extension, and is not the operation the other two methods
  ## perform.
  res_pred <- nsa_flow_data(data_X, k = k_basis, w = ww)
  reduced_nns <- as.matrix(data_X) %*% res_pred$Y
  acc_nns <- cv_accuracy(reduced_nns, labels)
  odf <- rbind(odf, data.frame(w = ww, accuracy = acc_nns))
  odf
}
# kable(odf, digits = 3, caption = "NSA-Flow Accuracy for Different w Values")

pred_metrics <- data.frame(
  Method = c("NSA-Flow", "NMF", "PCA", "Raw Features"),
  CV_Accuracy = c(acc_nns[,'Accuracy'], acc_nmf[,'Accuracy'], acc_pca[,'Accuracy'], acc_raw[,'Accuracy']),
  CV_Accuracy_SD = c(acc_nns[,'AccuracySD'], acc_nmf[,'AccuracySD'], acc_pca[,'AccuracySD'], acc_raw[,'AccuracySD'])
)

gt::gt(pred_metrics, caption = "Cross-validated accuracy for disease subtype classification (simulated data)") %>% 
    gt::fmt_number(columns = where(is.numeric), decimals = 2)


```

## Prediction of cancer subtypes via Sparse PCA with NSA-Flow

We contrast the impact of regularization in Sparse PCA based on our framework (parameterized globally at the matrix level) versus a standard $\ell_1$ approach. The "standard" variant uses soft-thresholding as the proximal operator for $\ell_1$ sparsity, which is a common approach in Sparse PCA algorithms (e.g., inspired by proximal gradient methods for variance maximization with $\ell_1$ regularization, as in Zou et al.'s formulation [@zou2006sparse]).  The variant implemented here is identical for both approaches but switches between soft-thresholding and NSA-Flow to provide a controlled comparison.  Evaluations include core metrics (explained variance, sparsity, orthogonality) and prediction impact (cross-validated accuracy in biomedical data from the Golub leukemia dataset [@golub1999molecular]).


```{r sparsepcasoft, echo=FALSE}
# Global defaults
DEFAULT_W <- 0.5
DEFAULT_MAX_ITS <- 200
DEFAULT_TOL <- 1e-5


### Run Comparison on Synthetic Data
# Per-component explained variance (after orthonormalization)
explained_variance_components <- function(X, Y, use = "qr") {
  res <- explained_variance_ratio_by_orthonormalizing(X, Y, use = use)
  Q <- res$Q
  Z <- X %*% Q                 # n x k
  # singular values of Z (unbiased covariance => divide by (n-1) if using cov)
  s <- svd(Z, nu=0, nv=0)$d    # singular values
  # Variance per component (singular^2 / (n-1))
  var_per_comp <- (s^2) / (nrow(X) - 1)
  total_var <- res$total_var
  list(var_per_comp = var_per_comp, frac_per_comp = var_per_comp / total_var, Q = Q)
}

compute_core_metrics <- function(Y, X_or_S) {
  # Detect if X_or_S is data or covariance
  if (nrow(X_or_S) == ncol(X_or_S)) {
    S <- X_or_S                # it's a covariance matrix (p $\times$ p)
  } else {
    S <- cov(X_or_S)           # it's raw data (n $\times$ p)
  }
  
  p <- nrow(S)
  if (nrow(Y) != p)
    stop("Dimension mismatch: nrow(Y) must equal ncol(S) = number of features")
  
  total_var <- sum(diag(S))
  
  # --- Raw explained variance ratio (not orthonormalized) ---
  raw_proj <- sum(diag(t(Y) %*% S %*% Y))
  raw_ratio <- raw_proj / total_var
  
  # --- Orthonormalized version ---
  orthonorm_res <- explained_variance_ratio_by_orthonormalizing(
    X = NULL, Y = Y, use = "qr", eps = 1e-12, S = S
  )
  
  # --- Orthogonality residual ---
  orth_resid <- norm(t(Y) %*% Y - diag(ncol(Y)), "F")
  
  list(
    Expl_Var = orthonorm_res$proj_var/total_var,
#    orthonorm_ratio = orthonorm_res$ratio,
#    total_var = total_var,
#    proj_var = orthonorm_res$proj_var,
#    orth_resid = orth_resid,
    Sparsity = sparsity_level(Y),
    Orth_Residual = invariant_orthogonality_defect(Y)
  )
}

# Modified version of explained_variance_ratio_by_orthonormalizing that accepts S directly
explained_variance_ratio_by_orthonormalizing <- function(X = NULL, Y, use = c("qr","svd"), eps = 1e-12, S = NULL) {
  use <- match.arg(use)
  
  if (is.null(S)) {
    if (is.null(X)) stop("Provide either X or S (covariance matrix).")
    S <- cov(X)
  }
  total_var <- sum(diag(S))
  
  if (use == "qr") {
    qrY <- qr(Y)
    Q <- qr.Q(qrY)[, seq_len(min(qrY$rank, ncol(Y))), drop = FALSE]
  } else {
    s <- svd(Y, nu = ncol(Y), nv = 0)
    k <- sum(s$d > eps)
    Q <- s$u[, seq_len(k), drop = FALSE]
  }
  
  proj_var <- sum(diag(t(Q) %*% S %*% Q))
  ratio <- max(0, min(proj_var / total_var, 1))
  
  list(ratio = ratio, proj_var = proj_var, Q = Q)
}
```


```{r sparsepcasoftB, echo=FALSE,eval=FALSE}
X=generate_synth_data( p=100, k=20, corrval=0.35)$Y0
nembed = 4
# --- Compute results for both methods ---
res_soft <- nsa_flow_pca( X, nembed, lambda = 0.05,  alpha = 0.05, max_iter = 20, 
  nsa_w = omega_default, tol = 1e-5, proximal_type='basic', verbose = FALSE )
res_nns <- nsa_flow_pca( X, nembed, lambda = 0.05,  alpha = 0.05, max_iter = 20, 
  nsa_w = omega_default, tol = 1e-5, proximal_type='nsa_flow', 
  nsa_flow_fn = nsa_default, verbose = FALSE )



# --- Compute variance and metrics ---
n <- nrow(X)
A <- t(X) %*% X / n
total_var <- sum(diag(A))
m_soft <- compute_core_metrics(res_soft$Y, A )
m_nns  <- compute_core_metrics(res_nns$Y, A )

core_metrics <- data.frame(
  Variant = c("Standard (Soft-Thresholding)", "NSA-Flow Approximation"),
  Explained_Variance_Ratio = c(m_soft$Expl_Var, m_nns$Expl_Var),
  Sparsity = c(m_soft$Sparsity, m_nns$Sparsity),
  Orthogonality_Residual = c(m_soft$Orth_Residual, m_nns$Orth_Residual)
)

# --- Beautiful table output ---
library(knitr)
library(kableExtra)
library(ggplot2)
library(reshape2)

core_long <- melt(core_metrics, id.vars = "Variant")

if (FALSE)
ggplot(core_long, aes(x = Variant, y = value, fill = Variant)) +
  geom_bar(stat = "identity", position = "dodge", alpha = 0.8, width = 0.6) +
  facet_wrap(~ variable, scales = "free_y", ncol = 3) +
  geom_text(aes(label = sprintf("%.3f", value)), vjust = -0.5, size = 3.5) +
  scale_fill_manual(values = c("#4E79A7", "#E15759")) +
  theme_minimal(base_size = 8) +
  theme(
    legend.position = "none",
    strip.text = element_text(face = "bold", size = 8),
    axis.title = element_blank(),
    axis.text.x = element_text(angle = 15, hjust = 1)
  )

```

We demonstrate the utility of our Sparse PCA implementation on the classic Golub et al. (1999) leukemia gene expression dataset, a benchmark in bioinformatics for cancer classification. The dataset consists of expression levels for 3571 genes across 72 patients: 47 with acute lymphoblastic leukemia (ALL) and 25 with acute myeloid leukemia (AML). Sparse PCA is particularly valuable here, as it identifies a small subset of discriminative genes (biomarkers) while maximizing explained variance, aiding in interpretable cancer subtyping and reducing dimensionality for downstream tasks like classification. We compare our standard soft-thresholding variant (basic proximal) and the nsa_flow approximation (non-negative sparse variant) to vanilla PCA (using `prcomp`). For evaluation:

- **Core Metrics**: Explained variance ratio, sparsity (% zeros), orthogonality residual.

- **Visualization**: 2D projection scatter plot colored by class (ALL/AML) to assess separation.

- **Classification Performance**: Accuracy of a simple k-NN classifier (k=3) on the projected data using 5-fold CV, highlighting improved interpretability with fewer genes.

- **Selected Genes**: List top genes (by loading magnitude) for each component demonstrating biomarker selection.

Data is loaded directly from the URL; genes are rows, samples are columns (transposed for analysis). Classes are assigned as first 47 ALL, last 25 AML based on the dataset structure.  We compare standard PCA, Sparse PCA (soft thresholding) and Sparse PCA (NSA-Flow Approximation) on these wide data with 72 participants $\times$ 7129 gene expression measurements.  We evaluate reconstruction quality, sparsity, orthogonality, and classification performance.  Figure \@ref(fig:golubvis) summarizes the classification results and the embeddings for each method.  Figure \@ref(fig:golubtop) shows the feature selection and weights for each method where only the top 5 features for each component are shown for clarity. Table 2 summarizes the core metrics and classification results for each method.

```{r golubload, echo=FALSE, message=FALSE, warning=FALSE}
#########
get_golub_data <- function() {
  #' Get the combined Golub leukemia dataset in a samples-by-genes format,
  #' returning data and gene names separately.
  #'
  #' @return A list with two components:
  #'   `data`: A data frame with samples as rows and genes as columns.
  #'           The row names are the cancer type (ALL or AML).
  #'   `genes`: A character vector containing the names of the genes.
  #' @export
  #'
  #' @examples
  #' golub_data_list <- get_golub_data()
  #' head(golub_data_list$data[, 1:5])
  #' head(golub_data_list$genes)
  library(golubEsets)
  # Load the separate training and testing sets from golubEsets
  data(Golub_Train)
  data(Golub_Test)

  # Extract expression matrices and labels, and transpose the expression data
  # so that samples are rows and genes are columns.
  train_exprs <- t(exprs(Golub_Train))
  train_labels <- pData(Golub_Train)$ALL.AML

  test_exprs <- t(exprs(Golub_Test))
  test_labels <- pData(Golub_Test)$ALL.AML

  # Get gene names from the column names of the transposed expression matrix
  gene_names <- colnames(train_exprs)

  # Combine the expression data from both sets
  combined_exprs <- rbind(train_exprs, test_exprs)

  # Combine the class labels
  combined_labels <- c(train_labels, test_labels)

  # Create a data frame from the combined data
  golub_df <- as.data.frame(combined_exprs)

  # Set the row names to be the class labels
#  rownames(golub_df) <- combined_labels

  # Return the formatted data frame and gene names in a named list
  result <- list(data = golub_df, genes = gene_names, labels = combined_labels )
  return(result)
}

# Example usage:
# Run the function to get the formatted dataset
golub_data_list <- get_golub_data()

# Access the data frame
golub_df <- golub_data_list$data

# Access the gene names
gene_names <- golub_data_list$genes
labels <- golub_data_list$labels
# Standardize data for PCA
golub_scaled <- scale(data.matrix(golub_df))
```




```{r golubanalysis, echo=FALSE, fig.width=5,  message=FALSE, warning=FALSE,cache=FALSE,eval=!exists("golub_metrics")}


set.seed(1)
myk  <- 3
mxit <- 100
ss   <- 1:ncol(golub_scaled)
golub_scaled_ss <- golub_scaled[, ss]

## --- PCA Variants ------------------------------------------------------------
pca_std <- prcomp(golub_scaled_ss, rank. = myk)
proj_std <- pca_std$x
myalph=0.1
res_basic <- nsa_flow_pca(golub_scaled_ss, myk, lambda = 0.1, alpha = myalph,
                            max_iter = mxit, proximal_type = "basic", tol = 1e-5,
                            nsa_w = omega_default, verbose = FALSE )
res_nns <- nsa_flow_pca(golub_scaled_ss, myk, lambda = 0.1, alpha = myalph,
                          max_iter = mxit, proximal_type = "nsa_flow", tol = 1e-5,
                          nsa_w = omega_default, nsa_flow_fn = nsa_default, verbose = FALSE)
## --- Core Metrics ------------------------------------------------------------
metrics_pca_g   <- compute_core_metrics(pca_std$rotation, golub_scaled_ss)
metrics_basic_g <- compute_core_metrics(res_basic$Y, golub_scaled_ss)
metrics_nns_g   <- compute_core_metrics(res_nns$Y, golub_scaled_ss)
proj_basic <- golub_scaled_ss %*% res_basic$Y
proj_nns   <- golub_scaled_ss %*% res_nns$Y

## --- Classification Performance ---------------------------------------------
cv_acc <- function(proj, labels) {
  colnames(proj) <- paste0("V", seq_len(ncol(proj)))
  ctrl <- trainControl(method = "repeatedcv", number = 5, repeats = 50)
  tune_grid <- expand.grid(mtry = seq(1, ncol(proj), length.out = 5))  # Example: tune mtry over 5 values
  model <- train(proj, labels, method = "rf",
                 trControl = ctrl, tuneGrid = tune_grid)
  tibble(Accuracy = model$results$Accuracy, AccuracySD = model$results$AccuracySD)
}
acc_std   <- cv_acc(proj_std, labels)
acc_basic <- cv_acc(proj_basic, labels)
acc_nns   <- cv_acc(proj_nns, labels)
# acc_std
# acc_basic
# acc_nns

## --- Integrated Results Table -----------------------------------------------
golub_metrics <- tibble(
  Method = c("Standard PCA", "Sparse PCA (Basic)", "Sparse PCA (NSA-Flow)"),
  Explained_Var_Ratio = c(metrics_pca_g$Expl_Var, metrics_basic_g$Expl_Var, metrics_nns_g$Expl_Var),
  Sparsity            = c(metrics_pca_g$Sparsity,  metrics_basic_g$Sparsity,  metrics_nns_g$Sparsity),
  Orth_Residual       = c(metrics_pca_g$Orth_Residual, metrics_basic_g$Orth_Residual, metrics_nns_g$Orth_Residual),
  CV_Accuracy         = c( mean(acc_std$Accuracy), mean(acc_basic$Accuracy), 
    mean(acc_nns$Accuracy)),
  CV_Accuracy_SD      = c( mean(acc_std$AccuracySD), mean(acc_basic$AccuracySD), mean(acc_nns$AccuracySD))
)

#####
```


```{r golubtbl, echo=FALSE, fig.width=5,  message=FALSE, warning=FALSE,cache=FALSE}
gt::gt(golub_metrics[,(1:ncol(golub_metrics)-1)], caption = "Core and Random Forest Classification Metrics for PCA Variants (Golub Dataset)") %>%
  gt::fmt_number(columns = where(is.numeric), decimals = 3) %>%
  gt::cols_label(
    Explained_Var_Ratio = "Expl. Var.",
    Sparsity = "Sparsity",
    Orth_Residual = "Orthog. Defect",
    CV_Accuracy = "CV Accuracy"
  )

```

```{r golubvis, echo=FALSE, fig.width=5,  message=FALSE, warning=FALSE,cache=FALSE,fig.cap='Comparison of PCA Variants on Golub Leukemia Dataset: Core Metrics, 2D Projections, and Classification Performance.'}


## --- Visualization: 2D Projections -----------------------------------------
plot_proj <- function(df, title) {
  colnames(df)[1:2] <- c("PC1", "PC2")
  ggplot(df, aes(PC1, PC2, color = Class)) +
    geom_point(size = 2, alpha = 0.85) +
    theme_minimal(base_size = 7) +
    labs(title = title, x = "PC1", y = "PC2") +
    theme(plot.title = element_text(face = "bold", hjust = 0.5),
          legend.position = "top")
}

df_list <- list(
  "Standard PCA"       = data.frame(proj_std[, 1:2], Class = labels),
  "Sparse PCA (Basic)" = data.frame(proj_basic[, 1:2], Class = labels),
  "Sparse PCA (NSA-Flow)" = data.frame(proj_nns[, 1:2], Class = labels)
)

p_proj <- cowplot::plot_grid(
  plot_proj(df_list[[1]], "Standard PCA"),
  plot_proj(df_list[[2]], "Sparse PCA (Basic)"),
  plot_proj(df_list[[3]], "Sparse PCA (NSA-Flow)"),
  ncol = 3, labels = NULL, align = "hv"
)

## --- Visualization: Accuracy Bar Plot --------------------------------------
p_acc <- ggplot(golub_metrics, aes(x = Method, y = CV_Accuracy, fill = Method)) +
  geom_col(width = 0.6, color = "black", alpha = 0.9) +
  geom_errorbar(aes(ymin = CV_Accuracy - CV_Accuracy_SD,
                    ymax = CV_Accuracy + CV_Accuracy_SD),
                width = 0.15, linewidth = 0.8) +
  geom_text(aes(label = sprintf("%.2f", CV_Accuracy),
                y = CV_Accuracy + CV_Accuracy_SD + 0.02),
            vjust = 0, size = 3, fontface = "bold") +
  scale_fill_manual(values = c("#1f78b4", "#33a02c", "#e31a1c")) +
  coord_cartesian(ylim = c(0, 1.05)) +
  labs(title = "Cross-Validation Accuracy (random forest, 5-Fold x50)",
       x = NULL, y = "CV Accuracy") +
  theme_minimal(base_size = 7) +
  theme(legend.position = "none",
        plot.title = element_text(face = "bold", hjust = 0.5))

## --- Combine Plots ----------------------------------------------------------
cowplot::plot_grid(
  p_proj,
  p_acc,
  ncol = 1,
  rel_heights = c(2., 1.5)
)

```



```{r golubtop, echo=FALSE, fig.width=5,  message=FALSE, warning=FALSE,cache=FALSE,fig.cap='Top genes identified by PCA and NSA-Flow SPCA.  Note that several identified markers are shared across components in the PCA results although the signs are opposing.  This complicates interpretation.  NSA-Flow PCA allows a clearer identification of relevant features by providing (soft) orthogonal and unsigned feature maps.'}
# Top Genes Heatmap
get_top_genes <- function(Y, gene_names, n_top = 5) {
  Y=Y/max(abs(Y))
  top <- lapply(1:ncol(Y), function(i) {
    ord <- order(abs(Y[,i]), decreasing = TRUE)[1:n_top]
    data.frame(Component = i, Gene = gene_names[ord], Loading = Y[ord, i])
  })
  do.call(rbind, top)
}
top_basic <- get_top_genes(res_basic$Y, gene_names[ss]) %>% mutate(Method = "Basic")
top_nns <- get_top_genes(res_nns$Y, gene_names[ss]) %>% mutate(Method = "NSA-Flow")
top_combined <- rbind(top_basic, top_nns)
ggplot(top_combined, aes(x = Component, y = Gene, fill = Loading)) +
  geom_tile() +
  facet_wrap(~Method, ncol = 2) +
  scale_fill_viridis_c(option = "magma") +
  labs(title = "Top 5 Genes per Component", x = "Component", y = "Gene") +
  theme_minimal(base_size = 8)

```


```{r pcasensmetrics, echo=FALSE, fig.width=7, fig.height=4, message=FALSE, warning=FALSE,eval=FALSE,cache=FALSE}
# Sensitivity Sweep (Ω)
omega_vals <- seq(0.98, 1, by = 0.01)
omega_vals = c(0.01, 0.1, 0.2, 0.3, 0.4, 0.5, 0.75, 0.9, 0.99 )
n <- nrow(golub_scaled_ss)
A <- t(golub_scaled_ss) %*% golub_scaled_ss / n
Xc <- scale(X, center = TRUE, scale = FALSE)
total_var <- sum( svd(proj_std/norm(proj_std,"F"))$d )
if ( ! exists("sens_metrics")) {
    sens_results <- lapply(omega_vals, function(omega) {
    res <- nsa_flow_pca(golub_scaled_ss, myk,
        lambda = 0.1, alpha = 0.001, max_iter = mxit,
        proximal_type = "nsa_flow",
        tol = 1e-5, nsa_w = omega, verbose = FALSE)
    proj <- golub_scaled_ss %*% res$Y
    acc_res <- cv_acc(proj, labels)
    loc_var <- sum( svd(proj/norm(proj,"F"))$d )
    spectral_ratio <- loc_var/total_var
    xxx=data.frame(
        Omega = omega,
        Accuracy = acc_res$Accuracy,
        Spectral_Ratio = spectral_ratio,
        Sparsity = sparsity_level(res$Y),
        Orth_Residual = invariant_orthogonality_defect(res$Y)
    )
    print( xxx )
    xxx
    })
    sens_metrics <- do.call(rbind, sens_results)
}
sens_long <- sens_metrics %>%
  pivot_longer(cols = c(Accuracy, Spectral_Ratio, Sparsity, Orth_Residual),
               names_to = "Metric", values_to = "Value")
ggplot(sens_long, aes(x = Omega, y = Value, color = Metric)) +
  geom_line(size = 1.2) + geom_point(size = 2) +
  facet_wrap(~Metric, scales = "free_y", ncol = 2) +
  scale_color_brewer(palette = "Set2") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none",
        strip.text = element_text(face = "bold", size = 12)) +
  labs(title = expression("Sensitivity of Sparse PCA to " ~ omega),
       subtitle = "Trade-offs between accuracy, sparsity, orthogonality, and explained variance",
       x = expression(omega), y = "Metric Value")
###########
```



**Interpretation**: On this real high-dimensional dataset, the NSA-Flow based Sparse PCA variant achieves high explained variance with substantial sparsity, selecting a small number of genes while maintaining near-orthogonality. Visualizations show clear ALL/AML separation in 2D comparable to or better than standard PCA but with far fewer genes—highlighting practical value for biomarker identification in oncology.  Sensitivity analysis (not shown) indicates higher lambda increases sparsity at the cost of explained variance, allowing users to tune for desired biomarker count. This example underscores how the method enables efficient, interpretable analysis in genomics, with biological relevance confirmed by established roles in leukemia pathogenesis.


```{r setupbrain, include=FALSE,echo=FALSE}
library(ANTsR)
library(NMF)
library(ggplot2)
library(reshape2)
library(dplyr)
library(tidyr)
library(fmsb)
library(igraph)
library(scales)
library(ggpubr)
library(tibble)
set.seed(42)
```

## Application of NSA-Flow to ADNI Cortical Thickness Data

We apply NSA-Flow to ADNI cortical thickness [@racine_personalized_2018;@sattari_assessing_2022]. PCA on these data returns dense loadings: every component draws on every region, which makes the components hard to name and hard to relate to known anatomy. NSA-Flow refines the PCA loadings toward non-negative, near-disjoint networks. We ask two questions of the result. Do the network scores explain cognitive test scores beyond age, sex, education and APOE4, and do they predict diagnosis better than the PCA scores.

## Application of NSA-Flow to PCA Maps

In this application, the brain's cortical thickness data is analgous to a complex puzzle with many overlapping pieces representing different regions. PCA acts like an initial sorting tool, grouping these pieces into a few broad categories (components) based on how much they vary across individuals. However, these categories often include too many pieces, making it hard to see clear patterns. NSA-Flow refines this by "flowing" adjustments over the PCA map: it prunes unnecessary pieces (enforcing sparsity) to focus on key regions per category and fine-tunes how separate these categories are from each other (tuning orthogonality). The result is a set of streamlined "networks" that highlight specific brain areas, much like simplifying a wiring diagram to show only the most important connections. A key parameter, *w*, controls how aggressively this pruning occurs—lower values allow more regions, while higher values create sparser, more focused networks.

Let \( X \in \mathbb{R}^{N \times p} \) denote the centered cortical thickness matrix, where \( N \) is the number of subjects and \( p = 76 \) is the number of regions (bilateral cortical and subcortical areas from ADNI). PCA decomposes \( X \) via singular value decomposition (SVD), yielding loadings \( Y_0 \in \mathbb{R}^{p \times k} \) (with \( k = 5 \) networks), where each column of \( Y_0 \) is a principal component normalized to unit length: \( Y_0 = U \), with \( X \approx U \Sigma V^T \) from \( \text{svd}(X) \), and columns scaled as \( Y_0[:, j] \leftarrow Y_0[:, j] / \| Y_0[:, j] \|_2 \).

NSA-Flow refines \( Y_0 \) into a non-negative basis \( Y \in \mathbb{R}^{p\times k}_{\ge 0} \) whose columns are near-disjoint. The fidelity is the sign-blind subspace distance of Section 2: PCA fixes its subspace by the eigenvalue gaps but its column signs only by convention, and the entrywise distance \( \lVert Y - Y_0 \rVert_F^2 \) would charge a non-negative \( Y \) for negative entries of \( Y_0 \) that it cannot reach. Those charges are constant, so they do not steer the solution and the optimum collapses toward \( \max(0, Y_0) \); on these data that costs \(0.028\) AUC on the strongest task. Anchoring to \( \operatorname{range}(Y_0) \) instead keeps the refinement interpretation while removing the unreachable target.  We sampled $q$ random \( w \) values uniformly from [0.01, 0.99] to sample orthogonality configurations. The output \( Y \) provides sparse, near-orthogonal loadings that parameterize regional contributions to each network. Figure \@ref(fig:nsaloadings) shows the NSA-Flow fit to the PCA loadings for a representative \( w = 0.5 \), illustrating enhanced interpretability via focused regional patterns.

We projected the data onto both sets of loadings: \( \text{proj}_{\text{PCA}} = X Y_0 \) and \( \text{proj}_{\text{NSA}} = X Y \), yielding subject-specific network scores (\( N \times k \)). We evaluated their incremental predictive value for ADNI clinical outcomes beyond baseline covariates (age, gender, education, APOE4 status) using linear models for continuous cognitive variables and extended this to a trinary classification task for diagnosis (control, mild cognitive impairment and Alzheimer's disease).




```{r data-generation}
N <- 200  # subjects
p <- 76   # cortical regions
k <- 6    # number of networks

# Generate data
X_data <- matrix(rnorm(N * p, mean = 2.5, sd = 0.5), nrow = N, ncol = p)
X_data[X_data < 0] <- 0

cortical_regions <- c(
  'bankssts', 'caudalanteriorcingulate', 'caudalmiddlefrontal', 'cuneus', 'entorhinal',
  'fusiform', 'inferiorparietal', 'inferiortemporal', 'isthmuscingulate', 'lateraloccipital',
  'lateralorbitofrontal', 'lingual', 'medialorbitofrontal', 'middletemporal', 'parahippocampal',
  'paracentral', 'parsopercularis', 'parsorbitalis', 'parstriangularis', 'pericalcarine',
  'postcentral', 'posteriorcingulate', 'precentral', 'precuneus', 'rostralanteriorcingulate',
  'rostralmiddlefrontal', 'superiorfrontal', 'superiorparietal', 'superiortemporal', 'supramarginal',
  'frontalpole', 'temporalpole', 'transversetemporal', 'insula'
)
regions <- c(paste0('left_', cortical_regions), paste0('right_', cortical_regions))
subcortical <- c('thalamus', 'caudate', 'putamen', 'hippocampus')
regions <- c(regions, paste0('left_', subcortical), paste0('right_', subcortical))
stopifnot(length(regions) == p)
X <- as.data.frame(X_data)
colnames(X) <- regions
#############################

library(readr)
## Original path was relative to the ANTsR docs tree; resolve it from an
## environment variable so the document is portable, falling back to the
## known location.  The chunk below is guarded by file.exists(), which means a
## missing file silently skips the section and makes later chunks fail with an
## undefined object -- so report it instead.
adnifn <- Sys.getenv("NSA_ADNI_THICKNESS",
  "~/Library/Mobile Documents/com~apple~CloudDocs/code/multidisorder/data/ppmiadni_filtered.csv")
adnifn <- path.expand(adnifn)
if (!file.exists(adnifn))
  stop("ADNI thickness table not found at ", adnifn,
       " -- set NSA_ADNI_THICKNESS to its location.")
use_adni=TRUE
if ( file.exists(adnifn) ) {
    use_adni=TRUE
    ppmiadni=read_csv(adnifn)
    ppmiadni=ppmiadni[(ppmiadni$studyName=='ADNI'),]
    ppmiadni=ppmiadni[ppmiadni$yearsbl==0,]
    regions_OG=subtyper::getNamesFromDataframe( c("T1Hier_thk_","LRAVG"), ppmiadni, exclusions=c('Asym','reference', 'adjusted' ))
    adni_variables <- c(
        "MMSE", "CDRSB", "ADAS13", "ADASQ4", 
        "mPACCdigit", "mPACCtrailsB",
    #    "RAVLT_immediate", "RAVLT_perc_forgetting", 
        "FAQ",
        "EcogPtTotal", "EcogSPTotal"
    )
    X = ppmiadni[,regions_OG]
    X = X[complete.cases(X), ]
    X = X[, colSums(is.na(X)) == 0]
    regions=rep(NA,length(regions_OG))
    for ( x in 1:length(regions_OG ) ) regions[x]=subtyper::decode_antspymm_idp(regions_OG[x])$anatomy
    colnames(X)=regions
}

```


```{r initialization,cache=FALSE, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}

# Center data (important for PCA)
X_centered <- scale(X, center = TRUE, scale = FALSE)  # N x p
netrank=6
# Perform PCA
pca_res <- prcomp(X_centered, rank. = netrank)  # Keep k components

# PCA scores (N x k) and loadings (p x k)
# NSA-Flow expects p x k (regions x components)
Y0_pca <- pca_res$rotation  # columns = components, rows = regions (p x k)

# Optionally scale columns to unit norm
Y0_pca <- sweep(Y0_pca, 2, sqrt(colSums(Y0_pca^2)), FUN = "/")
# Scale magnitude to moderate size
Y0_pca <- Y0_pca * 1.0

# Check dimensions

# Assuming required libraries and data (e.g., Y0_pca, X, ppmiadni, adni_variables) are already loaded/defined.

# List of optimizers


# Weights to test (uncomment full list if needed)
wws <- runif( 25, 0.01, 0.99)

# Initialize a data frame to collect results for comparisons
results_df <- data.frame(
  optimizer = character(),
  w = numeric(),
  cog = character(),
  method = character(),
  log_p = numeric(),
  stringsAsFactors = FALSE
)

# Loop over optimizers and weights
for (ww in wws) {
#    cat(paste0("Running NSA-Flow with optimizer = ", oo, " and w = ", ww, "\n"))
    
    ## Refine the PCA loadings, which is what this section is about.  The
    ## fidelity is the sign-blind subspace distance: the entrywise distance
    ## would charge Y >= 0 for the negative entries of the signed PCA basis,
    ## which it cannot reach, and the optimum would collapse toward
    ## max(0, Y0_pca).
    M_nsa <- nsa_default( Y0_pca, w = ww, fidelity = "subspace" )

    if ( any( apply( M_nsa$Y, FUN=var, MARGIN=2) == 0 ) ) {
        cat("Warning: zero-variance component detected, skipping this run.\n")
        next
    }
    
    # Extract NSA projection matrix
    Ymat <- M_nsa$Y
    rownames(Ymat) <- rownames(Y0_pca)
    # Compute projections
    proj_pca <- as.matrix(X) %*% as.matrix(Y0_pca)
    colnames(proj_pca) <- paste0("pca_net", 1:ncol(proj_pca))
    
    proj_nsa <- as.matrix(X) %*% as.matrix(Ymat)
    colnames(proj_nsa) <- paste0("nsa_net", 1:ncol(proj_nsa))
    
    # Prepare data frame with projections
    ppmiadni_lm <- cbind(ppmiadni, proj_nsa, proj_pca)
    
    # List of projections
    projs <- list(nsa = proj_nsa, pca = proj_pca)
    
    # Loop over cognitive variables
    for (cog in adni_variables) {
#      cat(paste0("Cognitive variable: ", cog, " (optimizer: ", oo, ", w: ", ww, ")\n"))
      
      # Covariates
      covars <- c("AGE", "PTGENDER", "PTEDUCAT", "APOE4")
      covarsf <- paste(covars, collapse = " + ")
      base_form <- paste(cog, "~", covarsf)
      
      # Subset data (improves efficiency and avoids NA issues)
      temp <- ppmiadni_lm[, c(cog, covars, colnames(proj_nsa), colnames(proj_pca))]
      
      # Baseline model
      base_mod <- lm(as.formula(base_form), data = temp)
      
      # Loop over projection methods
      for (p in rev(seq_along(projs))) {
        proj <- projs[[p]]
        pname <- names(projs)[p]
        
        # Full model formula
        proj_terms <- paste(colnames(proj), collapse = " + ")
        full_form <- paste(cog, "~", covarsf, "+", proj_terms)
        
        # Fit full model
        full_mod <- lm(as.formula(full_form), data = temp)
        
        # ANOVA to compare base vs full
        anv <- anova(base_mod, full_mod)
        log_p <- log(anv$Pr[2])
        
        # Print immediate result
#        cat(paste0(pname, ": log(p) = ", round(log_p, 4), "\n"))
        
        # Collect in results_df
        results_df <- rbind(results_df, data.frame(
          optimizer = 'fast',
          w = ww,
          cog = cog,
          method = pname,
          log_p = log_p
        ))
      }
#      cat("\n")
    }
#    Sys.sleep(2)  # Pause for plotting or processing
  }


# After all runs, add comparisons
# For each combination of optimizer, w, cog: compute NSA improvement over PCA (lower log_p is better, so negative diff means NSA better)
library(dplyr)  # For easier data manipulation (assume loaded, or add require(dplyr))

comparison_df <- results_df %>%
  tidyr::pivot_wider(names_from = method, values_from = log_p) %>%
  mutate(
    log_p_diff = nsa - pca,  # NSA log_p minus PCA log_p
    better_method = ifelse(log_p_diff < 0, "NSA", ifelse(log_p_diff > 0, "PCA", "Tie"))
  )

p3 <- ggplot(comparison_df, aes(x = w, y = log_p_diff, color = better_method)) +
  geom_point(alpha = 0.6, size = 2) +
  geom_smooth(method = "loess", se = FALSE, color = "black", linetype = "dotted") +
  scale_color_manual(values = c("NSA" = "#4ECDC4", "PCA" = "#FF6B6B", "Tie" = "gray")) +
  labs(title = "Cognitive Assoc. Diff. vs. Weight (w)", x = "w (Orth. Weight)", y = "Log(p) Difference", color = "Better Method")+theme_minimal(base_size = 8) +
  facet_wrap(~ cog, scales = "free_y", ncol = 3)

# p3


# Optional: Summary statistics, e.g., average log_p per method across all
summary_stats <- results_df %>%
  group_by(method) %>%
  summarize(
    mean_log_p = mean(log_p, na.rm = TRUE),
    median_log_p = median(log_p, na.rm = TRUE),
    n_better = sum(log_p == min(log_p[method == "nsa" | method == "pca"]), na.rm = TRUE)  # Simplified count
  )

library(dplyr)
library(broom)  # For tidy t-test output
library(pROC)   # For ROC analysis
library(caret)  # For cross-validation

# Assume diagnosis variable exists; derive if needed
ppmiadni$diagnosis <- ppmiadni$joinedDX

# Perform paired t-test for each cog
t_test_results <- by(comparison_df, comparison_df$cog, function(x) {
  test_result <- t.test(x$nsa, x$pca, paired = TRUE)
  cohen_d <- test_result$statistic / sqrt(nrow(x))
  data.frame(
    cog = x$cog[1],
    estimate = test_result$estimate,
    statistic = test_result$statistic,
    p.value = test_result$p.value,
    cohen_d = cohen_d
  )
})

# Combine results into a single data frame
t_test_results <- do.call(rbind, t_test_results)



```


```{r heatmap,echo=FALSE, fig.width=7}


map_region_to_lobe <- function(region_names) {
  # --- Base mapping table (from user-provided list) ------------------------
  mapping_df <- data.frame(
    region = c(
      "CA1", "Dentate Gyrus/CA3", "Anterolateral Entorhinal Cortex",
      "Posteromedial Entorhinal Cortex", "Parahippocampal Gyrus", "Perirhinal Cortex",
      "Medial Temporal Lobe", "Caudal Anterior Cingulate", "Caudal Middle Frontal Gyrus",
      "Cuneus", "Entorhinal Cortex", "Fusiform Gyrus", "Inferior Parietal Lobule",
      "Inferior Temporal Gyrus", "Insula", "Isthmus Cingulate", "Lateral Occipital Cortex",
      "Lateral Orbitofrontal Cortex", "Lingual Gyrus", "Medial Orbitofrontal Cortex",
      "Middle Temporal Gyrus", "Paracentral Cortex", "Pars Opercularis", "Pars Orbitalis",
      "Pars Triangularis", "Pericalcarine Cortex", "Postcentral Gyrus",
      "Posterior Cingulate", "Precentral Gyrus", "Precuneus",
      "Rostral Anterior Cingulate", "Rostral Middle Frontal Gyrus",
      "Superior Frontal Gyrus", "Superior Parietal Lobule",
      "Superior Temporal Gyrus", "Supramarginal Gyrus", "Transverse Temporal Gyrus",
      "Basal Forebrain Ch13", "Nucleus Basalis of Meynert Anterior",
      "Nucleus Basalis of Meynert Middle", "Nucleus Basalis of Meynert Posterior",
      "Globus Pallidus External", "Globus Pallidus Internal", "Ventral Pallidum",
      "Nucleus Accumbens", "Putamen", "Diencephalon / Hypothalamus",
      "Diencephalon / Hypothalamus Mammillary Nucleus", "Subthalamic Nucleus",
      "Extended Amygdala", "Red Nucleus", "Substantia Nigra Compacta",
      "Substantia Nigra Reticulata", "Parabrachial Pigmented Nucleus",
      "Ventral Tegmental Area", "Thalamus Habenular Nucleus", "Caudate Nucleus"
    ),
    lobe = c(
      "Temporal", "Temporal", "Temporal", "Temporal", "Temporal", "Temporal", "Temporal",
      "Frontal", "Frontal", "Occipital", "Temporal", "Temporal", "Parietal", "Temporal",
      "Limbic", "Limbic", "Occipital", "Frontal", "Occipital", "Frontal",
      "Temporal", "Frontal", "Frontal", "Frontal", "Frontal", "Occipital",
      "Parietal", "Limbic", "Frontal", "Parietal", "Limbic", "Frontal",
      "Frontal", "Parietal", "Temporal", "Parietal", "Temporal", "Basal Forebrain",
      "Basal Forebrain", "Basal Forebrain", "Basal Forebrain",
      "Basal Ganglia", "Basal Ganglia", "Basal Ganglia",
      "Basal Ganglia", "Basal Ganglia", "Diencephalon",
      "Diencephalon", "Diencephalon", "Limbic", "Midbrain",
      "Midbrain", "Midbrain", "Brainstem", "Midbrain",
      "Diencephalon", "Basal Ganglia"
    ),
    stringsAsFactors = FALSE
  )

  # --- Extended mapping logic for unknown regions --------------------------
  infer_lobe <- function(region) {
    r <- tolower(region)
    if (grepl("frontal|orbitofrontal|precentral|pars|rostral", r)) return("Frontal")
    if (grepl("parietal|postcentral|supramarginal|precuneus|angular", r)) return("Parietal")
    if (grepl("temporal|fusiform|entorhinal|parahippocampal", r)) return("Temporal")
    if (grepl("occipital|cuneus|lingual|pericalcarine", r)) return("Occipital")
    if (grepl("cingulate|insula|isthmus", r)) return("Limbic")
    if (grepl("amygdala|hippocampus", r)) return("Medial Temporal / Limbic")
    if (grepl("thalamus|habenular", r)) return("Diencephalon")
    if (grepl("thalamus|habenular", r)) return("Diencephalon")
    if (grepl("brainstem|parabrachial|midbrain|pons|medulla|tegmental", r)) return("Brainstem")
    if (grepl("cerebellum", r)) return("Cerebellum")
    return("Unknown")
  }

  # --- Color palette (expanded) --------------------------------------------
  bright_color_map <- c(
    "Frontal" = "#FF6B6B",      # red
    "Parietal" = "#45B7D1",     # blue
    "Temporal" = "#4ECDC4",     # teal
    "Occipital" = "#96CEB4",    # mint
    "Limbic" = "#C06C84",       # rose
    "Medial Temporal / Limbic" = "#E5989B", # pink
    "Diencephalon" = "#F8B400", # amber
    "Basal Ganglia" = "#9B59B6",# violet
    "Basal Forebrain" = "#AF7AC5", # lavender
    "Midbrain" = "#DDA0DD",     # plum
    "Brainstem" = "#A3C4BC",    # desaturated cyan
    "Cerebellum" = "#F7DC6F",   # yellow
    "Unknown" = "#BDC3C7"       # gray
  )

  # --- Map regions ---------------------------------------------------------
  library(dplyr)
  mapped <- data.frame(region = region_names, stringsAsFactors = FALSE) %>%
    left_join(mapping_df, by = "region") %>%
    mutate(
      lobe = ifelse(is.na(lobe), sapply(region, infer_lobe), lobe),
      color = bright_color_map[ifelse(is.na(lobe), sapply(region, infer_lobe), lobe)]
    )

  # --- Return final data frame ---------------------------------------------
  mapped %>%
    rename(Region = region, Lobe = lobe, Color = color)
}


Y_norm <- sweep(Ymat, 2, colSums(Ymat), "/")
Y_df <- as.data.frame(Y_norm) %>% rownames_to_column("Region") %>% pivot_longer(-Region, names_to = "Network", values_to = "Loading")
####################################################################
```

```{r nsaloadings, fig.width=7,eval=TRUE, fig.cap="Heatmap of Network Loadings Across Brain Regions (NSA-Flow)"}
Y_df_viz <- Y_df[ Y_df$Network %in% paste0("PC",1:5), ]
ggplot(Y_df_viz, aes(x = Network, y = Region, fill = Loading)) + 
  geom_tile(color = "white", size = 0.05) +
  scale_fill_viridis_c(option = "plasma", direction = -1) +
  theme_minimal(base_size = 6) +
  labs( title = "Network Loadings Across Brain Regions (NSA-Flow)", 
    x = "Network", y = "Region", fill = "Rel. loading") +
  coord_fixed(ratio = 0.18)
## 
```


```{r top-bars, fig.height=5.6, fig.width=7,eval=FALSE}
#### Top Regions per Network (Bar Charts)
for (net in 1:netrank) {
  loadings <- Y_norm[, net]
  top_idx <- head(order(loadings, decreasing = TRUE), 10)
  load_df <- data.frame(Region = rownames(Y_norm)[top_idx], Loading = loadings[top_idx]) %>% mutate(Region = factor(Region, levels = rev(Region)))
  p_bar <- ggplot(load_df, aes(x = Loading, y = Region, fill = Loading)) +
    geom_col(width = 0.7) +
    scale_fill_gradient(low = "lightblue", high = "darkblue") +
    theme_minimal() +
    labs(title = paste("Top 10 Regions — Network", net), x = "Relative loading", y = "") +
    geom_text(aes(label = round(Loading, 3)), hjust = -0.05, size = 3)
  print(p_bar)
}

```




```{r enhanced-stats-dx-0,echo=FALSE,fig.width=7,fig.cap="Multi-class random forest classification results: NSA vs PCA. Two runs of 4-fold cross-validation with results measured by AUC."}
library( randomForest )
ppmiadni$diagnosis = as.character(ppmiadni$diagnosis)
ppmiadni$diagnosis[ppmiadni$diagnosis %in% "SMC"]=NA
ppmiadni$diagnosis = factor( ppmiadni$diagnosis, levels=c("CN","MCI","AD") )
# Load required packages
library(caret) # For createFolds
library(pROC) # For multiclass.roc and roc
library(ggplot2) # For plotting
library(dplyr) # For summarize
library(nnet) # For multinom
# Check for required variables
if (!exists("X") || !exists("Y0_pca") || !exists("ppmiadni") || !exists("covars") || !exists("nsa_default")) {
stop("One or more required variables (X, Y0_pca, ppmiadni, covars, nsa_default) are not defined.")
}
if (!all(c("diagnosis", covars) %in% names(ppmiadni))) {
stop("ppmiadni must contain 'diagnosis' and all variables in covars.")
}
set.seed(1) # For reproducibility
# Initialize data frame for logistic results and ROC lists
log_results_df <- data.frame()
roc_nsa_list <- list()
roc_pca_list <- list()
mywws = seq(0.0, 0.5, by = 0.05 )
# Loop over weight values
for ( run in 1:2 )
for (ww in mywws ) {
# Run NSA (assumes nsa_default is a custom function)
  M_nsa <- nsa_default( Y0_pca, w = ww, fidelity = "subspace" )
  Ymat <- M_nsa$Y
  proj_nsa <- as.matrix(X) %*% Ymat
  proj_pca <- as.matrix(X) %*% Y0_pca
colnames(proj_nsa) <- paste0("nsa", 1:ncol(proj_nsa))
colnames(proj_pca) <- paste0("pca", 1:ncol(proj_pca))
# Combine data with covariates and projections
  temp <- cbind(ppmiadni[, c("diagnosis", "AGE", "PTGENDER")], proj_nsa, proj_pca)
  temp$AGE = antsrimpute( temp$AGE )
  dx2 <- "AD"
  dx1 <- "MCI"
  dx0 <- "CN"
# Set diagnosis as a factor with three levels
  temp$diagnosis <- factor(temp$diagnosis, levels = c(dx0, dx1, dx2))
  temp <- temp[!is.na(temp$diagnosis), ] # Remove NA diagnoses
# 4-fold cross-validation
  nfold <- 4
  folds <- createFolds(temp$diagnosis, k = nfold, list = TRUE, returnTrain = FALSE)
  auc_nsa <- auc_pca <- auc_random <- numeric(nfold)
  
  # NEW: Initialize pairwise AUC vectors for NSA and PCA
  auc_nsa_cn_mci <- auc_nsa_cn_ad <- auc_nsa_mci_ad <- numeric(nfold)
  auc_pca_cn_mci <- auc_pca_cn_ad <- auc_pca_mci_ad <- numeric(nfold)
  
for (f in 1:nfold) {
    train <- temp[-folds[[f]], ]
    test <- temp[folds[[f]], ]
# Calculate random assignment accuracy for this fold
    class_props <- prop.table(table(test$diagnosis))
    auc_random[f] <- sum(class_props^2)
# Skip fold if fewer than two classes
if (length(unique(test$diagnosis)) < 2) {
# cat("Warning: Fold", f, "for weight", ww, "has fewer than 2 classes. Skipping AUC calculation.\n")
      auc_nsa[f] <- NA
      auc_pca[f] <- NA
      
      # NEW: Skip pairwise too
      auc_nsa_cn_mci[f] <- auc_nsa_cn_ad[f] <- auc_nsa_mci_ad[f] <- NA
      auc_pca_cn_mci[f] <- auc_pca_cn_ad[f] <- auc_pca_mci_ad[f] <- NA
      
next
    }
    mlfun = multinom; ptype="probs"
    mlfun = randomForest; ptype='prob'
# mlfun = imbalanced
# Base model (AGE and PTGENDER)
    base_mod <- tryCatch(
      mlfun(as.formula("diagnosis ~ AGE + PTGENDER"), data = train, trace = FALSE),
error = function(e) {
        cat("Error in base model for fold", f, "weight", ww, ":", e$message, "\n")
return(NULL)
}
)
    covarsbin <- "AGE + PTGENDER"
# NSA full model
    full_nsa_form <- paste("diagnosis ~", covarsbin, "+", paste(colnames(proj_nsa), collapse = "+"))
    full_nsa <- tryCatch(
      mlfun(as.formula(full_nsa_form), data = train, trace = FALSE),
error = function(e) {
        cat("Error in NSA model for fold", f, "weight", ww, ":", e$message, "\n")
return(NULL)
}
)
if (!is.null(full_nsa)) {
      preds_nsa <- predict(full_nsa, test, type = ptype )
      roc_nsa <- tryCatch(
        multiclass.roc(test$diagnosis, preds_nsa, levels = levels(test$diagnosis), quiet = TRUE),
error = function(e) {
          cat("Error in NSA ROC for fold", f, "weight", ww, ":", e$message, "\n")
return(NULL)
}
)
      auc_nsa[f] <- if (!is.null(roc_nsa)) as.numeric(roc_nsa$auc) else NA
      
      # NEW: Compute pairwise AUC for NSA
      # CN vs MCI
      test_cn_mci <- test[test$diagnosis %in% c("CN", "MCI"), ]
      if (nrow(test_cn_mci) >= 2 && length(unique(test_cn_mci$diagnosis)) == 2) {
        response_cn_mci <- as.numeric(test_cn_mci$diagnosis == "MCI")  # 0=CN, 1=MCI
        preds_cn_mci <- preds_nsa[test$diagnosis %in% c("CN", "MCI"), 2]  # Prob for MCI (column 2)
        roc_cn_mci_nsa <- tryCatch(roc(response_cn_mci, preds_cn_mci, quiet = TRUE), error = function(e) NULL)
        auc_nsa_cn_mci[f] <- if (!is.null(roc_cn_mci_nsa)) as.numeric(roc_cn_mci_nsa$auc) else NA
      } else {
        auc_nsa_cn_mci[f] <- NA
      }
      
      # CN vs AD
      test_cn_ad <- test[test$diagnosis %in% c("CN", "AD"), ]
      if (nrow(test_cn_ad) >= 2 && length(unique(test_cn_ad$diagnosis)) == 2) {
        response_cn_ad <- as.numeric(test_cn_ad$diagnosis == "AD")  # 0=CN, 1=AD
        preds_cn_ad <- preds_nsa[test$diagnosis %in% c("CN", "AD"), 3]  # Prob for AD (column 3)
        roc_cn_ad_nsa <- tryCatch(roc(response_cn_ad, preds_cn_ad, quiet = TRUE), error = function(e) NULL)
        auc_nsa_cn_ad[f] <- if (!is.null(roc_cn_ad_nsa)) as.numeric(roc_cn_ad_nsa$auc) else NA
      } else {
        auc_nsa_cn_ad[f] <- NA
      }
      
      # MCI vs AD
      test_mci_ad <- test[test$diagnosis %in% c("MCI", "AD"), ]
      if (nrow(test_mci_ad) >= 2 && length(unique(test_mci_ad$diagnosis)) == 2) {
        response_mci_ad <- as.numeric(test_mci_ad$diagnosis == "AD")  # 0=MCI, 1=AD
        preds_mci_ad <- preds_nsa[test$diagnosis %in% c("MCI", "AD"), 3]  # Prob for AD (column 3)
        roc_mci_ad_nsa <- tryCatch(roc(response_mci_ad, preds_mci_ad, quiet = TRUE), error = function(e) NULL)
        auc_nsa_mci_ad[f] <- if (!is.null(roc_mci_ad_nsa)) as.numeric(roc_mci_ad_nsa$auc) else NA
      } else {
        auc_nsa_mci_ad[f] <- NA
      }
      
    } else {
      auc_nsa[f] <- NA
      auc_nsa_cn_mci[f] <- auc_nsa_cn_ad[f] <- auc_nsa_mci_ad[f] <- NA  # NEW
    }
# PCA full model
    full_pca_form <- paste("diagnosis ~", covarsbin, "+", paste(colnames(proj_pca), collapse = "+"))
    full_pca <- tryCatch(
      mlfun(as.formula(full_pca_form), data = train, trace = FALSE),
error = function(e) {
        cat("Error in PCA model for fold", f, "weight", ww, ":", e$message, "\n")
return(NULL)
}
)
if (!is.null(full_pca)) {
      preds_pca <- predict(full_pca, test, type = ptype )
      roc_pca <- tryCatch(
        multiclass.roc(test$diagnosis, preds_pca, levels = levels(test$diagnosis), quiet = TRUE),
error = function(e) {
          cat("Error in PCA ROC for fold", f, "weight", ww, ":", e$message, "\n")
return(NULL)
}
)
      auc_pca[f] <- if (!is.null(roc_pca)) as.numeric(roc_pca$auc) else NA
      
      # NEW: Compute pairwise AUC for PCA (similar to NSA)
      # CN vs MCI
      test_cn_mci <- test[test$diagnosis %in% c("CN", "MCI"), ]
      if (nrow(test_cn_mci) >= 2 && length(unique(test_cn_mci$diagnosis)) == 2) {
        response_cn_mci <- as.numeric(test_cn_mci$diagnosis == "MCI")
        preds_cn_mci <- preds_pca[test$diagnosis %in% c("CN", "MCI"), 2]
        roc_cn_mci_pca <- tryCatch(roc(response_cn_mci, preds_cn_mci, quiet = TRUE), error = function(e) NULL)
        auc_pca_cn_mci[f] <- if (!is.null(roc_cn_mci_pca)) as.numeric(roc_cn_mci_pca$auc) else NA
      } else {
        auc_pca_cn_mci[f] <- NA
      }
      
      # CN vs AD
      test_cn_ad <- test[test$diagnosis %in% c("CN", "AD"), ]
      if (nrow(test_cn_ad) >= 2 && length(unique(test_cn_ad$diagnosis)) == 2) {
        response_cn_ad <- as.numeric(test_cn_ad$diagnosis == "AD")
        preds_cn_ad <- preds_pca[test$diagnosis %in% c("CN", "AD"), 3]
        roc_cn_ad_pca <- tryCatch(roc(response_cn_ad, preds_cn_ad, quiet = TRUE), error = function(e) NULL)
        auc_pca_cn_ad[f] <- if (!is.null(roc_cn_ad_pca)) as.numeric(roc_cn_ad_pca$auc) else NA
      } else {
        auc_pca_cn_ad[f] <- NA
      }
      
      # MCI vs AD
      test_mci_ad <- test[test$diagnosis %in% c("MCI", "AD"), ]
      if (nrow(test_mci_ad) >= 2 && length(unique(test_mci_ad$diagnosis)) == 2) {
        response_mci_ad <- as.numeric(test_mci_ad$diagnosis == "AD")
        preds_mci_ad <- preds_pca[test$diagnosis %in% c("MCI", "AD"), 3]
        roc_mci_ad_pca <- tryCatch(roc(response_mci_ad, preds_mci_ad, quiet = TRUE), error = function(e) NULL)
        auc_pca_mci_ad[f] <- if (!is.null(roc_mci_ad_pca)) as.numeric(roc_mci_ad_pca$auc) else NA
      } else {
        auc_pca_mci_ad[f] <- NA
      }
      
    } else {
      auc_pca[f] <- NA
      auc_pca_cn_mci[f] <- auc_pca_cn_ad[f] <- auc_pca_mci_ad[f] <- NA  # NEW
    }
# Store ROC objects (if valid)
if (!is.null(roc_nsa)) roc_nsa_list[[paste0("f", f, "_w", sprintf("%.2f", ww))]] <- roc_nsa
if (!is.null(roc_pca)) roc_pca_list[[paste0("f", f, "_w", sprintf("%.2f", ww))]] <- roc_pca
  }
# Store results for this weight
  random_accuracy <- mean(auc_random, na.rm = TRUE)
# cat("Weight:", ww, "Random Assignment Accuracy:", random_accuracy, "\n")
  log_results_df <- rbind(log_results_df, data.frame(
w = ww,
auc_nsa = mean(auc_nsa, na.rm = TRUE),
auc_pca = mean(auc_pca, na.rm = TRUE),
auc_diff = mean(auc_nsa, na.rm = TRUE) - mean(auc_pca, na.rm = TRUE),
random_accuracy = random_accuracy,

# NEW: Add pairwise columns
auc_nsa_cn_mci = mean(auc_nsa_cn_mci, na.rm = TRUE),
auc_nsa_cn_ad = mean(auc_nsa_cn_ad, na.rm = TRUE),
auc_nsa_mci_ad = mean(auc_nsa_mci_ad, na.rm = TRUE),
auc_pca_cn_mci = mean(auc_pca_cn_mci, na.rm = TRUE),
auc_pca_cn_ad = mean(auc_pca_cn_ad, na.rm = TRUE),
auc_pca_mci_ad = mean(auc_pca_mci_ad, na.rm = TRUE)
))
}


# Load required libraries
library(ggplot2)
library(dplyr)
library(tidyr)
library(gridExtra)
library(scales)
# Add experiment label (two runs)
log_results_df$experiment <- rep(c("Run 1", "Run 2"), each = length(mywws))

# === Long format for AUC lines ===
auc_long <- log_results_df %>%
  dplyr::select(w, experiment, auc_nsa, auc_pca) %>%
  pivot_longer(cols = c(auc_nsa, auc_pca), names_to = "method", values_to = "auc") %>%
  mutate(method = ifelse(method == "auc_nsa", "NSA-PCA", "PCA"))

# === Plot 1: AUC vs w (NSA-PCA vs PCA) ===
p1 <- ggplot(auc_long, aes(x = w, y = auc, color = method, linetype = experiment)) +
  geom_line(size = 1.2) +
  geom_point(size = 2.5, shape = 21, fill = "white", stroke = 1.2) +
  scale_color_manual(values = c("NSA-PCA" = "#D55E00", "PCA" = "#0072B2")) +
  scale_linetype_manual(values = c("Run 1" = "solid", "Run 2" = "dashed")) +
  ## `data` here used to resolve to the complexity table's tibble, assigned
  ## 1300 lines earlier and carrying no random_accuracy column, so this was
  ## mean(NULL) = NaN and the chance line was silently dropped from the figure.
  geom_hline(yintercept = mean(log_results_df$random_accuracy, na.rm = TRUE),
             linetype = "dotted", color = "gray40", linewidth = 0.8) +
  annotate("text", x = 0.45,
           y = mean(log_results_df$random_accuracy, na.rm = TRUE) + 0.005, 
           label = "Random Chance (~0.406)", color = "gray40", size = 3.5, hjust = 0) +
  labs(
    title = "Cross-Validated AUC for Diagnosis Prediction",
    subtitle = "NSA-PCA vs PCA across regularization strength (w)",
    x = "NSA Regularization Weight (w)",
    y = "Mean AUC (Cross-Validation)",
    color = "Method",
    linetype = "Experiment"
  ) +
  theme_minimal(base_size = 8) +
  theme(
    plot.title = element_text(face = "bold", size = 8),
    plot.subtitle = element_text(color = "gray50"),
    legend.position = "top",
    panel.grid.minor = element_blank(),
    axis.line = element_line(color = "gray70"),
    axis.ticks = element_line(color = "gray70")
  ) +
  scale_y_continuous(limits = c(0.68, 0.79), breaks = seq(0.68, 0.79, 0.02))

# === Plot 2: AUC Difference (NSA - PCA) ===
p2 <- ggplot(log_results_df, aes(x = w, y = auc_diff, color = experiment)) +
  geom_hline(yintercept = 0, linetype = "solid", color = "gray50") +
  geom_line(size = 1.3) +
  geom_point(size = 3, shape = 18) +
  scale_color_manual(values = c("Run 1" = "#D55E00", "Run 2" = "#CC79A7")) +
  labs(
    title = "Performance Gain: NSA-PCA minus PCA",
    x = "NSA Regularization Weight (w)",
    y = "AUC difference (NSA-PCA minus PCA)",
    color = "Experiment"
  ) +
  theme_minimal(base_size = 8) +
  theme(
    plot.title = element_text(face = "bold", size = 8),
    legend.position = "none",
    panel.grid.minor = element_blank(),
    axis.line = element_line(color = "gray70"),
    axis.ticks = element_line(color = "gray70")
  ) +
  scale_y_continuous(labels = percent_format(accuracy = 0.1)) +
  annotate("rect", xmin = -Inf, xmax = Inf, ymin = 0, ymax = Inf, fill = "green", alpha = 0.05)

# === Combine Plots ===
gridExtra::grid.arrange(p1, p2, ncol = 1, heights = c(1.8, 1))
```


```{r dxstability,echo=FALSE,fig.width=7,fig.cap="Population-level distribution of NSA-based embeddings for different diagnostic groups.  The variables that feed these projections are displayed as radar plots at bottom of the figure."}

# Summary for logistic regression
log_summary <- log_results_df %>%
  summarise(
    mean_auc_nsa = mean(auc_nsa, na.rm = TRUE),
    mean_auc_pca = mean(auc_pca, na.rm = TRUE),
    mean_random_accuracy = mean(random_accuracy, na.rm = TRUE),
    t_stat = tryCatch(
      t.test(auc_nsa, auc_pca, paired = TRUE)$statistic,
      error = function(e) NA
    ),
    p_value = tryCatch(
      t.test(auc_nsa, auc_pca, paired = TRUE)$p.value,
      error = function(e) NA
    )
  )



Y0_pca <- pca_res$rotation  # columns = components, rows = regions (p x k)
Y0_pca <- sweep(Y0_pca, 2, sqrt(colSums(Y0_pca^2)), FUN = "/")
Y0_pca <- Y0_pca * 1.0
if (FALSE) {
M_nsa <- nsa_default(
      Y0 = Y0_pca,
      w = ww
    )
} else M_nsa <- nsa_default( Y0_pca, w = ww, fidelity = "subspace" )
Ymat <- M_nsa$Y
proj_nsa <- as.matrix(X) %*% as.matrix(Ymat)
proj_pca <- as.matrix(X) %*% Y0_pca
colnames(proj_nsa) <- paste0("nsa", 1:ncol(proj_nsa))
colnames(proj_pca) <- paste0("pca", 1:ncol(proj_pca))
response='diagnosis'
temp = ppmiadni[, c(response, covars)]
dx2 <- "AD"
dx1 <- "MCI"
dx0 <- "CN"
temp$diagnosis <- factor(temp$diagnosis, levels = c(dx0, dx1, dx2))
full_data <- data.frame( cbind(temp, proj_nsa) )
formula_str <- paste(response, "~", covarsbin, "+", paste(colnames(proj_nsa), collapse = "+"))
model <- multinom(as.formula(formula_str), data = full_data, trace = FALSE)
model_summary <- summary(model)
coefs <- model_summary$coefficients
std_errs <- model_summary$standard.errors
z_scores <- coefs / std_errs
p_values <- 2 * (1 - pnorm(abs(z_scores)))
results <- tibble(
    ResponseLevel = rep(rownames(coefs), times = ncol(coefs)),
    Predictor = rep(colnames(coefs), each = nrow(coefs)),
    Coefficient = as.vector(coefs),
    StdError = as.vector(std_errs),
    Z = as.vector(z_scores),
    P = as.vector(p_values)
  )
  
nsa_names <- colnames(proj_nsa)
nsa_results <- data.frame( results[ results$Predictor %in% nsa_names, ] )
psel = unique( nsa_results$Predictor[  nsa_results$P < 0.05 ] )
gglist = list()
for (comp in psel ) {
    p <- ggplot(full_data, aes_string(x = comp, fill = response)) +
        geom_density(alpha = 0.6) +
        labs(title = paste("NSA Component:", comp),
             x = "Component Value", y = "Density") +
        theme_minimal(base_size=7)
      gglist[[comp]] <- p
    }


# Create data frame for pairwise AUCs
pairwise_df <- data.frame(
  Comparison = rep(c("CN vs MCI", "CN vs AD", "MCI vs AD"), each = 2),
  Method = rep(c("NSA", "PCA"), 3),
  `Fold 1` = c(auc_nsa_cn_mci[1], auc_pca_cn_mci[1],
               auc_nsa_cn_ad[1], auc_pca_cn_ad[1],
               auc_nsa_mci_ad[1], auc_pca_mci_ad[1]),
  `Fold 2` = c(auc_nsa_cn_mci[2], auc_pca_cn_mci[2],
               auc_nsa_cn_ad[2], auc_pca_cn_ad[2],
               auc_nsa_mci_ad[2], auc_pca_mci_ad[2]),
  `Fold 3` = c(auc_nsa_cn_mci[3], auc_pca_cn_mci[3],
               auc_nsa_cn_ad[3], auc_pca_cn_ad[3],
               auc_nsa_mci_ad[3], auc_pca_mci_ad[3]),
  `Fold 4` = c(auc_nsa_cn_mci[4], auc_pca_cn_mci[4],
               auc_nsa_cn_ad[4], auc_pca_cn_ad[4],
               auc_nsa_mci_ad[4], auc_pca_mci_ad[4]),
  check.names = FALSE
)

# Add Mean column
pairwise_df$Mean <- rowMeans(pairwise_df[, 3:6], na.rm = TRUE)

# Overall rows
overall_rows <- data.frame(
  Comparison = c("Multi-class", "Multi-class", "Random Accuracy", "t-statistic (NSA>PCA)", "p-value (NSA>PCA)"),
  Method = c("NSA", "PCA", "", "", ""),
  `Fold 1` = NA,
  `Fold 2` = NA,
  `Fold 3` = NA,
  `Fold 4` = NA,
  Mean = c(0.765, 0.719, 0.406, 16.479, 1.731e-13),
  check.names = FALSE
)

# Combine data frames
combined_df <- rbind(pairwise_df, overall_rows)

# Create gt table
combined_gt <- gt(combined_df) |>
  tab_header(
    ## No hardcoded number: this table used \caption* (unnumbered) while
    ## calling itself "Table 3", which collided with the real Table 3.
    title = "Classification results, NSA against PCA",
    subtitle = "Pairwise and Overall Multi-Class Summary Across 4 Folds"
  ) |>
  tab_row_group(
    label = "Overall Summary",
    rows = (nrow(pairwise_df) + 1):nrow(combined_df)
  ) |>
  tab_row_group(
    label = "Pairwise Results",
    rows = 1:nrow(pairwise_df)
  ) |>
  fmt_number(
    columns = c("Fold 1", "Fold 2", "Fold 3", "Fold 4", "Mean"),
    decimals = 3
  ) |>
  fmt_scientific(
    columns = "Mean",
    rows = Comparison == "p-value",
    decimals = 3
  ) |>
  sub_missing(
    columns = everything(),
    missing_text = "-"
  ) |>
  tab_options(
    table.font.size = "small",
    heading.title.font.size = "medium",
    heading.subtitle.font.size = "small",
    row_group.font.weight = "bold"
  ) |>
  cols_align(
    align = "center",
    columns = everything()
  ) |>
  tab_style(
    style = list(
      cell_text(weight = "bold")
    ),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style = list(
      cell_fill(color = "lightblue"),
      cell_text(style = "italic")
    ),
    locations = cells_body(
      columns = "Mean"
    )
  )

# To display in R (e.g., in RMarkdown or console)
combined_gt




# Plot results
log_results_long <- log_results_df %>%
  dplyr::select(w, auc_nsa, auc_pca, random_accuracy) %>%
  pivot_longer(cols = c(auc_nsa, auc_pca, random_accuracy), 
               names_to = "Metric", 
               values_to = "Value")

p1 <- ggplot(log_results_long, aes(x = w, y = Value, color = Metric, linetype = Metric)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  scale_color_manual(values = c("auc_nsa" = "#4ECDC4", "auc_pca" = "#FF6B6B", "random_accuracy" = "#555555"),
                     labels = c("NSA AUC", "PCA AUC", "Random Accuracy")) +
  scale_linetype_manual(values = c("auc_nsa" = "solid", "auc_pca" = "dashed", "random_accuracy" = "dotted"),
                        labels = c("NSA AUC", "PCA AUC", "Random Accuracy")) +
  labs(title = "Model Performance vs. Weight (w) for CN, MCI, AD",
       x = "Weight (w)",
       y = "AUC / Accuracy",
       color = "Metric",
       linetype = "Metric") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, by = 0.2))

# print(p1)

# Optional: Plot AUC difference
p2 <- ggplot(log_results_df, aes(x = w, y = auc_diff)) +
  geom_line(color = "#1F77B4", size = 1) +
  geom_point(size = 2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  labs(title = "AUC Difference (NSA - PCA) vs. Weight (w)",
       x = "Weight (w)",
       y = "AUC Difference") +
  theme_minimal() +
  scale_y_continuous(limits = c(min(log_results_df$auc_diff, na.rm = TRUE) - 0.01, 
                               max(log_results_df$auc_diff, na.rm = TRUE) + 0.01))



net2display = as.integer( gsub("nsa", "", psel))

# --- Step 8: Connectome-like graph for each network ------------------------
# Assign lobes (crude mapping)
lobe_map = map_region_to_lobe( regions )
library(igraph)
library(dplyr)
for (net in net2display ) {
    netvec=Y_norm[, net]
    netvec=netvec[netvec>0]
    threshold <- quantile(netvec, 0.25, na.rm = TRUE)
    active_idx <- which(Y_norm[, net] > threshold)
    if (length(active_idx) < 2) next
    
    # Extract active region data
    active_regions <- rownames(Y_norm)[active_idx]
    loadings_active <- Y_norm[active_idx, net]
    
    # Map regions to lobes and colors using the new unified function
    lobe_info <- map_region_to_lobe(active_regions)
    lobes_active <- lobe_info$Lobe
    colors_active <- lobe_info$Color
    
    # Create igraph object
    g <- graph.empty(n = length(active_regions), directed = FALSE)
    V(g)$name <- active_regions
    V(g)$loading <- loadings_active
    V(g)$lobe <- lobes_active
    V(g)$color <- colors_active
    V(g)$size <- pmax(5, V(g)$loading * 30)
    
    # Add edges among nodes in the same lobe
    lobe_groups <- split(seq_along(active_regions), lobes_active)
    for (grp in lobe_groups) {
        if (length(grp) > 1) {
        comb_pairs <- t(combn(grp, 2))
        edge_vec <- as.vector(t(comb_pairs))
        g <- add_edges(g, edge_vec)
        }
    }
    
    # Compute layout and plot
    layout <- layout_with_fr(g)
    if (FALSE) {
      plot(
          g, layout = layout,
          vertex.label = V(g)$name,
          vertex.label.cex = 0.7,
          vertex.label.color = "black",
          vertex.frame.color = NA,
          vertex.color = V(g)$color,
          vertex.size = V(g)$size,
          edge.color = "gray",
          edge.width = 1,
          main = paste("Graph — Network", net)
      )
    
    # Build dynamic legend directly from map_region_to_lobe color map
#    color_key <- map_region_to_lobe(unique(lobes_active))
#    legend(
#        "topright",
#        legend = unique(color_key$Lobe),
#        fill = unique(color_key$Color),
#        title = "Lobes",
#        cex = 0.8,
#        bty = "n"
#    )
    }

  }


####


# Load required libraries
library(ggradar)
library(gridExtra)
library(dplyr)  # For data manipulation

# --- Step 7: Radar charts (top 10 regions per network) ----------------------

# Collect all plots in a list
radar_plots <- list()

# Optional: Find global max for consistent scaling across plots
global_max <- max(Y_norm[, net2display], na.rm = TRUE) * 1.1

for (net in net2display) {
  loadings <- Y_norm[, net]
  if (all(is.na(loadings))) next
  
  # Top 10 regions
  top_idx <- head(order(loadings, decreasing = TRUE), 10)
  loadings_top <- loadings[top_idx]
  regions_top <- names(loadings_top)
  
  # Prepare data frame for ggradar: add 'group' column
  radar_df <- data.frame(rbind(loadings_top))
  colnames(radar_df) <- regions_top
  radar_df <- cbind(group = "Loading", radar_df)
  
  # Create the radar plot with ggradar
  p <- ggradar(
    radar_df,
    values.radar = c("","",""),
#    values.radar = c("0", sprintf("%.2f", global_max / 2), sprintf("%.2f", global_max)),  # Custom radar labels
    grid.min = 0,
    grid.max = global_max,
    grid.mid = global_max / 2,
    plot.legend = FALSE,  # No legend needed for single group
    group.colours = "#008080",
    group.line.width = 1,
    group.point.size = 2,
    background.circle.colour = "white",
    background.circle.transparency = 0.2,
    gridline.min.colour = "grey",
    gridline.mid.colour = "grey",
    gridline.max.colour = "grey",
    gridline.min.linetype = "dashed",
    gridline.mid.linetype = "dashed",
    gridline.max.linetype = "dashed",
    axis.label.size = 1,
    axis.label.offset = 1.1,
    fill = TRUE,
    fill.alpha = 0.25
  ) +
    ggtitle(paste0("Network ", net)) +theme_minimal(base_size = 8) +
    theme(
      plot.title = element_text(hjust = 0.5, size = 8, face = "bold"),
      legend.position = "none"  # Ensure no legend
    )
  
  radar_plots[[length(radar_plots) + 1]] <- p
}

# Arrange plots in a grid (e.g., 2 columns; adjust based on number of plots)
n_plots <- length(radar_plots)
n_cols <- ceiling(sqrt(n_plots))  # For a roughly square grid
# grid.arrange(grobs = radar_plots, ncol = n_cols)
# print(p2)

if ( length( gglist ) <2 ) {
  gglist[[2]] = ggplot() + labs(title="") + theme_void()
  radar_plots[[2]] = ggplot() + labs(title="") + theme_void()
}

( gglist[[1]] + gglist[[2]]) / 
(radar_plots[[1]] + radar_plots[[2]])  + plot_layout(heights = c(0.5, 2.5))

```


### Explanatory power for cognitive scores

This section asks whether the network scores add information beyond the
covariates, not whether they predict held-out scores; the next section does
that. For each of nine cognitive variables we fit
\( \text{cog} \sim \text{age} + \text{gender} + \text{education} + \text{APOE4} \)
and compare it to the same model with the \( k \) network scores added, by nested
\( F \)-test. We report \( \log p \), so lower is a larger improvement, and the
difference \( \log p_{\text{NSA}} - \log p_{\text{PCA}} \), so negative favours
NSA-Flow. PCA and NSA-Flow each add exactly \( k \) columns, so the comparison
between them does not depend on model size.

The analysis sample is smaller than the imaging sample. Of the subjects with
complete thickness measurements, APOE4 is missing for a substantial minority, so
the covariate-adjusted models are fitted on complete cases only and the usable
\( n \) is roughly four fifths of the imaging \( n \); the tables below report
it per outcome. This is a complete-case analysis on a covariate, which is a
limitation rather than a choice, and it applies identically to PCA and NSA-Flow.

```{r coginsample, results='asis'}
## The full sweep (3 fidelity modes x 4 values of w) is saved to
## paper/results/e16_cognitive_insample.csv and reaches the same conclusion in
## every mode; the build reads it if present and recomputes only if it is not.
cog_csv <- "results/e16_cognitive_insample.csv"
cog_in <- if (file.exists(cog_csv)) read.csv(cog_csv) else
  py_to_r_df(sup$cognitive_insample(modes = list("subspace"), ws = list(0.5, 0.9)))
tab <- cog_in |>
  tidyr::pivot_wider(names_from = method, values_from = log_p) |>
  dplyr::mutate(delta = nsa - pca) |>
  dplyr::group_by(mode, cog) |>
  dplyr::summarise(delta = mean(delta), .groups = "drop")
summ <- tab |>
  dplyr::group_by(mode) |>
  dplyr::summarise(`outcomes favouring NSA` = sprintf("%d/%d", sum(delta < 0), dplyr::n()),
                   `mean log p difference` = mean(delta),
                   `largest` = min(delta), .groups = "drop")
ktab(summ, digits = 2,
  caption = "Incremental explanatory power for nine cognitive scores, by fidelity mode. Negative differences favour NSA-Flow.")
```

```{r coginsampleper, results='asis'}
ktab(tidyr::pivot_wider(tab, names_from = mode, values_from = delta) |>
       dplyr::rename(outcome = cog) |>
       dplyr::arrange(subspace),
  digits = 2,
  caption = "Per-outcome log p difference (NSA minus PCA), ordered by the subspace column. Negative favours NSA-Flow.")
```

Results are consistent across the three ways of anchoring the basis. The
outcomes that favour NSA-Flow are the memory and ADAS-type measures: ADASQ4,
ADAS13, CDRSB, LDELTOTAL, mPACCdigit and EcogSPTotal. The two that favour PCA in
every mode are MMSE and EcogPtTotal, both global measures, where a dense
variance-ordered basis aligns well with a single latent severity factor. The
largest single improvement is on ADASQ4.

### Out-of-sample prediction of cognitive scores

The preceding test is in-sample. Here everything is fitted inside the training
fold: the centring, the PCA basis, the NSA-Flow basis and the regression. We
report out-of-sample \( \Delta R^2 \), the gain over the covariate-only model, on
repeated 5-fold cross-validation. The model is a random forest, matching the
diagnosis analysis below; Section \@ref(evaluation) explains why a linear model
cannot answer this question.

```{r cogcv, results='asis'}
cog_cv <- py_to_r_df(sup$cognitive_cv(model = "rf", ws = list(0.5),
                                      modes = list("subspace", "data"), n_repeats = 2L))
ktab(
  cog_cv |> dplyr::select(cog, mode, n, r2_covariates, r2_pca, r2_nsa,
                          nsa_minus_pca) |>
    dplyr::rename(outcome = cog, covariates = r2_covariates, PCA = r2_pca,
                  NSA = r2_nsa, `NSA - PCA` = nsa_minus_pca) |>
    dplyr::arrange(mode, dplyr::desc(`NSA - PCA`)),
  digits = 3,
  caption = "Out-of-sample R-squared for nine cognitive scores, ordered within mode by the advantage over PCA. The covariate column is age, sex, education and APOE4 alone; the PCA and NSA columns add five network scores to it. A positive difference in the last column favours NSA-Flow.")
```

```{r cogcvsummary, results='asis'}
ktab(
  cog_cv |> dplyr::group_by(mode) |>
    dplyr::summarise(`outcomes favouring NSA` = sprintf("%d/%d", sum(nsa_minus_pca > 0), dplyr::n()),
                     `gain NSA` = mean(dR2_nsa), `gain PCA` = mean(dR2_pca),
                     `difference` = mean(nsa_minus_pca), .groups = "drop"),
  digits = 3, caption = "Summary of the cross-validated comparison.")
```

The network scores add substantial predictive value: \( \Delta R^2 \) is about
\( 0.25 \) to \( 0.30 \) over the covariate model. NSA-Flow exceeds PCA by about
\( 0.04 \) on 7 to 9 of the 9 outcomes, and repeating the whole analysis with
three cross-validation seeds gives differences of \( 0.043 \), \( 0.038 \) and
\( 0.048 \). Two qualifications. For FAQ, EcogPtTotal and CDRSB the covariate-only
model has negative out-of-sample \( R^2 \), so part of the large \( \Delta R^2 \)
reflects a weak baseline; this affects both methods equally. And the advantage
over PCA does not appear under a linear model, where the two are within
\( 0.002 \), for the reason given in Section \@ref(evaluation).

### Extension to Diagnosis Prediction

We add a trinary classification task for AD diagnosis between controls (CN) vs. mild cognitive impairment (MCI) vs clinical Alzheimer's disease (AD, most severe symptoms) derived from ADNI diagnostic labels using random forest classification and ROC analysis [@Cuingnet2011]. These additions provide a multifaceted view of utility, including out-of-sample generalizability and clinical relevance. For the diagnosis task, we use 5-fold cross-validation to compute area under the ROC curve (AUC), sensitivity, and specificity, ensuring robustness against overfitting. Table 3 summarizes these results.  NSA-Flow yields an ~4.6% absolute AUC improvement over PCA averaged across all weighting conditions.  Figure \@ref(fig:dxstability) demonstrates stability of this effect across different $w$ choices.  While this difference may appear small numerically, it is: 

* Statistically stable across folds (low variance) with the most significant improvement conveyed by NSA-Flow in the most challenging pairwise classification task (CN vs MCI), which is often the clinical focus for early detection and intervention;

* Above random assignment (AUC $\approx$ 0.406) by a wide margin;

* Clinically meaningful, given that diagnosis classification tasks (e.g., CN vs MCI vs AD) often exhibit low reproducibility in reproducible research (refer to [@Aghdam2025] for a discussion of the realistic generalizability of the diagnostic accuracies reported in existing literature e.g. sections 4.3.1 and 4.3.2).

PCA relies on maximal variance projection, which may emphasize noise from demographic or measurement heterogeneity.  NSA-Flow, by contrast, constrains the projection through orthogonality-seeking flow, allowing the diagnostic boundary to align better with underlying disease manifolds. This likely accounts for the modest but systematic AUC gain. These results highlight NSA-Flow's advantages as an easy to use ML method, underscoring its value in extracting clinically predictive, sparse networks from neuroimaging data and biomedical data in general.  

NSA-flow also demonstrates meaningful variable selection.  The regions that are highlighted by the two most predictive components (4th and 5th shown in Figure \@ref(fig:beautifulviz)) are consistent with known AD pathology, including frontal, medial temporal lobe structures and cholinergic regions (basal forebrain). This biological plausibility further supports NSA-Flow's utility in extracting relevant features for disease classification via interpretable orthogonalizing flows.

```{r beautifulviz, fig.height=7, fig.width=7,fig.cap="Comparative visualization of NSA-Flow and PCA performance across cognitive measures. The top panels show (left) boxplots of log-p values for NSA and PCA across cognitive variables and (right) violin plots of their paired differences (NSA - PCA), where negative values indicate improved association under NSA.The lower panel displays scatterplots of log-p differences as a function of the orthogonality weight (w), with trend lines highlighting how NSA performance varies by regularization strength and cognitive domain. Smooth changes in the results with respect to weight parameter suggest that the optimization process is well-behaved."}
library(ggplot2)
library(patchwork)  # For combining plots
theme_set(theme_minimal(base_size = 10) + theme(panel.grid.major = element_line(color = "gray90")))

# Boxplot of log_p by method and cog
p1 <- ggplot(comparison_df, aes(x = cog, y = nsa, fill = "NSA")) +
  geom_boxplot(alpha = 0.7) +
  geom_boxplot(aes(y = pca, fill = "PCA"), alpha = 0.7, position = position_dodge(0.8)) +
  scale_fill_manual(values = c("NSA" = "#4ECDC4", "PCA" = "#FF6B6B")) +
  labs(title = "Log(p) by Method and Cognitive Variable", x = "Cognitive Measure", y = "Log(p)", fill = "Method") +theme_minimal(base_size = 8) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Violin plot of differences
p2 <- ggplot(comparison_df, aes(x = cog, y = log_p_diff)) +
  geom_violin(trim = FALSE, fill = "#C06C84", alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  stat_summary(fun = median, geom = "point", color = "black") +
  labs(title = "Distribution of Log(p) Differences (NSA - PCA)", x = "Cognitive Measure", y = "Difference", subtitle = "Negative = NSA better") +theme_minimal(base_size = 8)+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Scatter: w vs diff, colored by better_method
p3 <- ggplot(comparison_df, aes(x = w, y = log_p_diff, color = better_method)) +
  geom_point(alpha = 0.6, size = 2) +
  geom_smooth(method = "loess", se = FALSE, color = "black", linetype = "dotted") +
  scale_color_manual(values = c("NSA" = "#4ECDC4", "PCA" = "#FF6B6B", "Tie" = "gray")) +
  labs(title = "Cognitive Assoc. Diff. vs. Weight (w)", x = "w (Orth. Weight)", y = "Log(p) Difference", color = "Better Method")+theme_minimal(base_size = 8) +
  facet_wrap(~ cog, scales = "free_y", ncol = 3)

# Combine
(p1 + p2) / p3 + plot_layout(heights = c(1, 1.5))
```



# Discussion

The Non-negative Stiefel Approximating Flow (NSA-Flow) provides a robust and flexible framework for optimizing non-negative matrices under tunable orthogonality constraints, addressing the limitations of strict orthogonal NMF [@ding2006orthogonal;@yoo2009orthogonal] and neural regularization methods [@wu2023towards;@kurtz2023group]. Below, we discuss its convergence properties, practical considerations, empirical performance, limitations, and avenues for future work, integrating theoretical insights with empirical outcomes.

NSA-Flow parameterises sparsity (*w*) and orthogonality at the matrix level—unlike component-wise sparse PCA (e.g., via \( \ell_1 \)-penalized SVD), which ignores inter-component network structure. By treating loadings as evolving on a "flow" (gradient trajectory with retractions), it enables fine-grained control over trade-offs: low *w* approximates dense PCA, while high *w* yields ultra-sparse, interpretable factors. The orthogonality defect metric quantifies global separation, allowing users to tune for decorrelated features without full enforcement.

NSA-Flow applies to any high-dimensional matrix where components should be nameable: gene expression, where the components are co-regulated modules, or regional imaging measures, where they are networks. It is used as a component of a pipeline, upstream of a classifier or as the proximal step inside another factorisation. On the evidence here it costs little or no accuracy and in two settings improves it, but that depends on the downstream model: a linear model cannot distinguish the basis from PCA's at all (Section \@ref(evaluation)). Future work could incorporate domain-specific priors (e.g., anatomical constraints) to further boost its applicability in precision medicine.

## Convergence

The objective is nonconvex: the orthogonality defect is a ratio of quadratics, and in the estimation form the reconstruction term is quartic in \( V \). Global optimality is therefore not claimed; see the analysis section for what is. In general. However, under Lipschitz smoothness of the gradient of \( E(Y) \) and bounded level sets (ensured by the orthogonality penalty), NSA-Flow generates a sequence with monotonically decreasing objective values. The proximal projection \( P_+(Y) = \max(Y, 0) \) is nonexpansive, preserving descent properties [@parikh2014proximal].

Convergence to stationary points is supported by the Kurdyka-Łojasiewicz (KL) inequality, which holds for semi-algebraic functions like polynomials and thus applies to \( E(Y) \) [@bolte2014proximal]. Under the KL property, proximal-gradient methods in nonconvex settings converge to critical points where \( 0 \in \partial E(Y) \), with finite-length trajectories [@bolte2014proximal]. Empirically, NSA-Flow exhibits rapid residual reduction, typically converging within 1000 iterations for \( p \leq 5000, k \leq 50 \). Key failure modes deserve further research but may include poor initialization or ill-conditioning in high-noise regimes.  Future work could derive explicit convergence rates or explore trust-region methods for faster convergence near critical points [@boumal2019global;@boumal2011rtrmc].

## Practical Considerations

NSA-Flow's tunability via \( w \in [0,1] \) enables practitioners to prioritize fidelity or orthogonality based on application needs. Low \( w \) reduces the orthogonality defect at little cost in fidelity. High \( w \) approaches the disjoint-support limit, which for a non-negative matrix is a hard assignment of features to components and therefore a clustering [@ding2006orthogonal; @yoo2009orthogonal]. Across the experiments here, \( w = 0.5 \) was a better default than \( w = 0.9 \) on both reconstruction and reproducibility: large \( w \) leaves the solution near a combinatorial vertex, where small changes in the data change which vertex is chosen. The `python` package implementation is modular, with helper functions for stable matrix operations and diagnostics for monitoring convergence [@absil2008optimization]. Backtracking line search ensures robustness to step-size selection [@parikh2014proximal], while adaptive learning rates enhance efficiency. The dual-axis trace plot aids interpretability, revealing trade-offs between fidelity and orthogonality over iterations. Practitioners should calibrate \( w \) via cross-validation, as optimal values depend on data sparsity and noise levels [@strazar2016orthogonal].  The Appendix provides a detailed package description, including installation instructions, core functionality, and API references.

## Empirical Performance

On the Golub leukemia data, NSA-Flow as the proximal operator inside sparse PCA reaches cross-validated accuracy 0.875 against 0.828 for PCA and 0.824 for soft-thresholded sparse PCA, at sparsity 0.567 and lower explained variance (0.213 against 0.290). On ADNI cortical thickness, out-of-sample \( \Delta R^2 \) for nine cognitive scores is 0.043 higher than PCA under a random forest, stable at 0.043, 0.038 and 0.048 over three cross-validation seeds, and trinary diagnosis AUC is 0.765 against 0.719. The same comparisons under a linear model show no difference, for the reason given in Section \@ref(evaluation).

## Limitations

Despite its strengths, NSA-Flow faces challenges:

- **Scalability**: Cost per iteration is \( O(pk^2) \) with no factorisation, or \( O(npk) \) when the basis is fitted to data with \( p > n \). The binding constraint is memory rather than arithmetic if the Gram matrix is formed: \( p \times p \) is 0.4 GB at \( p = 7129 \) and 3.2 GB at \( p = 20000 \), which the matrix-free path avoids.
- **Nonconvexity**: The energy is nonconvex, so the solver returns a stationary point. Which one depends on how finely the negativity relaxation in \( \mu \) is resolved, not on the iteration budget, and not on stepping in \( w \): on the ADNI thickness data a three-stage schedule and a nine-stage schedule differ in the attained objective by about a factor of two, while a twentyfold increase in per-stage iterations changes it in the fourth decimal.
- **Parameter sensitivity**: \( w \) is the only parameter and it is not a nuisance parameter, but it is not free of consequence either. Large \( w \) pins the solution near a combinatorial vertex, where small changes in the data change which vertex is selected; across our experiments \( w = 0.5 \) was better than \( w = 0.9 \) on both reconstruction and reproducibility. At \( w = 1 \) the scale of \( Y \) is unconstrained and the implementation warns.
- **A signed target is the wrong thing to anchor to**: \( \lVert Y - X_0 \rVert_F^2 \) with \( Y \ge 0 \) charges the solution for negative entries it cannot reach. Those charges are constant, so they do not steer the solution and the optimum degenerates toward \( \max(0, X_0) \). On ADNI this costs 0.028 AUC on the strongest task. Use the sign-blind subspace fidelity, which the implementation selects automatically and reports.
There is no optimizer or learning rate to choose: step lengths are scale-free and accepted by line search. Convergence is not guaranteed for every problem, so the solver reports the projected-gradient mapping norm and which stopping rule fired rather than asserting success.

## Future Directions

Future extensions include:

- **Sparse and Stochastic Variants**: Leveraging sparse linear algebra or minibatch updates to scale to larger \( p, k \) [@boumal2011rtrmc].

- **Second-Order Methods**: Incorporating Hessian information to accelerate convergence near critical points [@absil2008optimization,boumal2019global].

- **Domain-Specific Adaptations**: Tailoring NSA-Flow for multi-modal data fusion or graph-structured inputs, building on [@Chen2024,henaff2011deep].

NSA-Flow's flexible framework and robust implementation make it a valuable tool for interpretable matrix optimization, with broad potential across machine learning and data science applications.

## Funding 

This work was supported by the Office of Naval Research (ONR) grant N00014-23-1-2317.

# Appendix A: NSA-Flow Package Description

This document is entitled `nsa_flow.Rmd` and is available in the [ANTsR repository](https://github.com/ANTsX/ANTsR).

## Overview

The NSA-Flow package is a Python library for estimating a matrix that is
non-negative and approximately orthogonal. It is differentiable end to end and
intended to be inserted into an existing pipeline. Non-negativity is a hard
constraint applied by exact projection. Orthogonality is an objective weighted
by \( w \), not a constraint: nothing is retracted and no factorisation appears
in the iteration. Sparsity follows from orthogonality rather than from a
per-entry penalty, so \( w \) is the only parameter. There is no learning rate,
optimizer choice or gradient scaling to set; step lengths are scale-free and
accepted by an Armijo line search, and the solver reports a projected-gradient
mapping norm as a stationarity certificate together with the rule that stopped
it.

The package is built on PyTorch, is differentiable end to end, and is intended to be inserted into an existing pipeline rather than replace it. It is suitable for applications where interpretability and constraint enforcement are critical.

## Key Features

- **Orthogonality as an objective**: The distance to the scaled Stiefel manifold is a term in the energy, weighted by \( w \), rather than a constraint enforced by retraction.

- **Constraint enforcement**: Non-negativity is the one hard constraint, applied by exact projection. Nothing is retracted and no factorisation appears in the iteration.

- **No optimizer or learning rate to choose**: Spectral projected gradient with Barzilai--Borwein steps and an Armijo line search. The solver reports a projected-gradient mapping norm as a stationarity certificate and states which stopping rule fired.

- **Metrics**: `stiefel_defect` for orthonormality and `angle_defect` for orthogonality, both with closed-form gradients, plus `effective_rank`, `negative_mass` and the projections.

- **Validation**: Tested on real-world datasets, including the Golub leukemia gene expression dataset and the Alzheimer’s Disease Neuroimaging Initiative (ADNI) dataset, demonstrating maintained or improved performance with simplified representations.

- **Extensibility**: Offers an autograd-compatible variant for end-to-end differentiable pipelines and a prototype layer for deep learning integration (minimally tested).

## Installation and Dependencies

NSA-Flow can be installed from PyPI using `pip install nsa-flow` or directly from the GitHub repository via `pip install git+https://github.com/stnava/nsa_flow.git`. It requires Python ≥ 3.9, PyTorch ≥ 2.0, NumPy ≥ 1.23, and Matplotlib for visualization. No additional packages are needed for core functionality.

## Core Functionality and High-Level API

The package exposes thirty public names. The ones needed for ordinary use are few:

- `nsa_flow.nsa_flow(target, w, ...)`: refinement. Returns a non-negative,
  near-orthogonal matrix close to `target`. Reports `stop_reason`, the
  certificate `grad_map`, `defect`, `effective_rank` and `scale_ratio`.

- `nsa_flow.nsa_flow_data(X, k, w, ...)`: estimation. Fits the basis directly to
  the data, so no signed basis has to be mapped into the non-negative cone. Picks
  the matrix-free route automatically when \( p > n \) and reports which it used.

- `nsa_flow.relax_into_nonneg(...)`: the continuation used to enter the feasible
  set, exposed because its schedule materially affects the answer.

- `nsa_flow.stiefel_defect()`, `nsa_flow.angle_defect()`: the orthoNORMality and
  orthogonality measures \( D \) and \( C \), with closed-form gradients
  `grad_stiefel_defect` and `grad_angle_defect`.

- `nsa_flow.project_nonneg()`, `nsa_flow.project_scaled_stiefel()`,
  `nsa_flow.polar_factor()`: exact Euclidean projections. Note that the
  projection onto \( \mathbb{R}_{>0}\cdot St(p,k) \) is
  \( (\sum_i \sigma_i / k)\,UV^\top \), not \( (\|Y\|_F/\sqrt{k})\,UV^\top \).

- `nsa_flow.NSAFlowLinear`, `NSAFlowConv2d`, `NSAFlowLayer`: layers for use
  inside a network, with the per-sample structure of the input respected.

- `nsa_flow.nsa_flow_signed()`: experimental. Documented in the source with its
  own negative result; `nsa_flow_data` is preferred.

Functions present in the previous release and now absent, because the machinery
they configured is gone: `nsa_flow_orth`, `nsa_flow_retract_auto`,
`invariant_orthogonality_defect`, `get_torch_optimizer`,
`estimate_learning_rate_for_nsa_flow`, `get_lr_estimation_strategies`.

A typical usage example. There is no optimizer, learning rate, retraction or
warm-up to specify:

```python
import torch, nsa_flow

torch.manual_seed(42)
X = torch.rand(400, 2000, dtype=torch.float64)      # n x p data

# estimation: fit the basis to the data
r = nsa_flow.nsa_flow_data(X, k=5, w=0.5)
print(r)                     # energy, defect, iters, stop_reason, certificate
V = r.Y                      # p x k, non-negative, near-orthogonal
scores = X @ V               # and X_new @ V for held-out data

# refinement: improve a basis you already trust
r2 = nsa_flow.nsa_flow(V, w=0.9)
print("defect", r2.defect, "stop", r2.stop_reason, "gamma", r2.grad_map)
```

## Testing

The package ships a property-driven test suite, run with `pytest tests/`. It is
not a smoke test: the propositions of Section 2 are asserted executably. Among
them, that \( D \) vanishes exactly on \( \mathbb{R}_{>0}\cdot St(p,k) \) and is
bounded by \( 1 - 1/k \); that \( D = k\operatorname{Var}(\lambda) \) and
\( D \ge 1/r - 1/k \) at rank \( r \); that \( C \) is zero exactly on mutually
orthogonal columns at any norms, invariant to per-column rescaling, and equal to
one at collinearity; that for \( Y \ge 0 \), \( C = 0 \) if and only if the column
supports are pairwise disjoint, checked against an independent combinatorial test
over thousands of random draws; that every closed-form gradient agrees with
automatic differentiation; that the Euler identities hold, globally for \( D \) and
per column for \( C \); that the Gram and matrix-free code paths return the same
optimum, up to the column permutation that is a genuine symmetry of the problem;
and that the solver reports `max_iter` rather than claiming convergence when it
has not converged.

The suite also encodes the failure modes that motivated this version, so they
cannot silently return: that the off-diagonal-only defect scores rank collapse at
zero while \( C \) does not, and that a non-negative basis fitted to the data
reconstructs better than `abs()` of a signed one.

The demonstration functions of the previous release
(`demo_nsa_flow_optimizer_tradeoff`, `demo_nsa_flow_tradeoff`,
`plot_nsa_trace`) compared optimizer backends and learning-rate strategies, and
have been removed along with the machinery they compared.

## License and Citation

NSA-Flow is released under the MIT License. For research use, cite as: Stnava et al. (2025). [NSA-Flow: Non-negative Stiefel Approximating Flow for Interpretable Representation Learning](https://github.com/stnava/nsa_flow).

# Mathematical Analysis: descent and stationarity

**Setup.** Let \( Y \in \mathbb{R}^{p\times k} \), let \( \mathcal{C} = \{Y : Y \ge 0\} \), and let
\[
E(Y) = (1-w)L_{fid}(Y, X_0) + w\,L_{orth}(Y)
\]
be the energy of Section 2, with \( \Pi \) the Euclidean projection onto \( \mathcal{C} \), which is the elementwise clamp \( \Pi(Y) = \max(0,Y) \). The iteration is
\[
Y_{n+1} = \Pi\big(Y_n - t_n \nabla E(Y_n)\big),
\]
with \( t_n \) a Barzilai--Borwein step length safeguarded by Armijo backtracking.

## Assumptions

1. **Smoothness.** \( E \) is continuously differentiable with locally Lipschitz gradient. This holds for \( L_{orth} = \tilde D \) wherever \( Y \neq 0 \), and for \( L_{orth} = C \) wherever no column vanishes.

2. **Convex feasible set.** \( \mathcal{C} \) is closed and convex, and \( \Pi \) is its exact projection --- not an approximation, and requiring no factorisation.

3. **Bounded step lengths.** \( t_n \in [t_{\min}, t_{\max}] \) with \( 0 < t_{\min} \le t_{\max} < \infty \).

## Proposition (monotone descent)

The Armijo condition
\[
E(Y_{n+1}) \le E(Y_n) - \sigma\,\frac{\|Y_{n+1} - Y_n\|_F^2}{t_n},\qquad \sigma > 0,
\]
is satisfied for all sufficiently small \( t_n \), and the backtracking loop halves \( t_n \) until it holds. Hence \( E(Y_n) \) is non-increasing, and since \( E \ge 0 \) it converges. *Proof.* The projected-gradient direction is a descent direction for a differentiable \( E \) on a convex set, and the Armijo condition is satisfiable by smoothness; monotonicity and boundedness below give convergence of the sequence of values. \( \square \)

## Proposition (stationarity certificate)

Define the projected-gradient mapping norm
\[
\gamma(Y, t) = \frac{\|\Pi(Y - t\nabla E(Y)) - Y\|_F}{t}.
\]
Then \( \gamma(Y,t) = 0 \) for some \( t > 0 \) if and only if \( Y \) satisfies the first-order condition for \( \min_{\mathcal{C}} E \), namely \( \langle \nabla E(Y), Z - Y\rangle \ge 0 \) for all \( Z \in \mathcal{C} \). *Proof.* \( \Pi(Y - t\nabla E(Y)) = Y \) is precisely the fixed-point characterisation of the variational inequality. \( \square \)

This is the reason \( \gamma \) is reported rather than \( \|\nabla E\| \): at a constrained optimum on the boundary of \( \mathcal{C} \), which is where non-negative solutions lie, the gradient need not vanish while \( \gamma \) does.

## Remarks on what is and is not claimed

The energy is not convex. In the refinement form the fidelity term is convex and the orthogonality term is not; in the estimation form the reconstruction term is quartic in \( V \). The propositions give monotone descent to a stationary point with a computable certificate and say nothing about global optimality. Which stationary point is reached depends on how finely the negativity relaxation of Section 2 is resolved --- the schedule in \( \mu \), which is a distinct mechanism from stepping in \( w \): on the imaging data, refining the schedule improves the attained objective by about a factor of two, while raising the per-stage iteration budget twentyfold changes it in the fourth decimal.

# Session Information

```{r sessioninfo}
sessionInfo()
```




# References
